---
format:
  revealjs:
    slide-level: 2
    slide-number: true
    preview-links: auto
    footer: "kerasnip v0.1.0.900"
    mermaid:
      theme: neutral
    width: 1280
    height: 720
    margin: 0.05
    min-scale: 0.2
    max-scale: 2.0
execute:
  freeze: auto
  cache: true
---

```{r}
#| label: madrid-2-2026-cache-results
#| include: false
#| message: false
#| warning: false

# Compute results once, cache them on disk, and reuse across EN/ES renders.
#
# Quarto executes this document with working directory = `presentations/madrid-2-2026/`,
# so cache files should live in `presentations/madrid-2-2026/results/` (i.e. `results/` here).
results_dir <- "results"
old_results_dirs <- c(
    file.path("..", "..", "madrid-2026", "results"),
    file.path("madrid-2026", "results")
)
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

# Migration:
# - Before the folder move, cached results lived at `madrid-2026/results/`.
# - Older versions could also write nested `madrid-2026/results/` inside the project dir.
for (old_results_dir in old_results_dirs) {
    if (!dir.exists(old_results_dir)) {
        next
    }

    old_rds <- list.files(old_results_dir, pattern = "\\.rds$", full.names = TRUE)
    if (length(old_rds) > 0) {
        file.copy(old_rds, results_dir, overwrite = FALSE)
    }
}

diamonds_rds <- file.path(results_dir, "diamonds_res.rds")
diamonds_final_rds <- file.path(results_dir, "diamonds_final.rds")
attrition_rds <- file.path(results_dir, "attrition_res.rds")
attrition_final_rds <- file.path(results_dir, "attrition_final.rds")

set.seed(20260203)

can_run <- function(...) {
    all(vapply(list(...), requireNamespace, logical(1), quietly = TRUE))
}

if (can_run("tidymodels", "kerasnip", "keras3")) {
    suppressPackageStartupMessages({
        library(tidymodels)
        library(kerasnip)
        library(keras3)
    })

    data(diamonds, package = "ggplot2")

    input_block <- function(model, input_shape) {
        keras_model_sequential(input_shape = input_shape)
    }

    dense_block <- function(model, units = 128, dropout = 0.0) {
        model |>
            layer_dense(units = units, activation = "relu") |>
            layer_dropout(rate = dropout)
    }

    output_block <- function(model) {
        model |> layer_dense(units = 1)
    }

    create_keras_sequential_spec(
        model_name = "diamond_mlp",
        layer_blocks = list(
            input = input_block,
            body = dense_block,
            output = output_block
        ),
        mode = "regression"
    )

    diamonds_split <- initial_split(diamonds, strata = price)
    diamonds_train <- training(diamonds_split)

    diamonds_rec <- recipe(price ~ ., data = diamonds_train) |>
        step_dummy(all_nominal_predictors()) |>
        step_zv(all_predictors()) |>
        step_normalize(all_numeric_predictors())

    spec <- diamond_mlp(
        num_body = tune(),
        body_units = tune(),
        body_dropout = tune(),
        fit_epochs = 5
    ) |>
        set_engine("keras")

    wf <- workflow() |>
        add_recipe(diamonds_rec) |>
        add_model(spec)

    params <- extract_parameter_set_dials(wf) |>
        update(
            num_body = dials::num_terms(c(1, 3)),
            body_units = dials::hidden_units(c(32, 128)),
            body_dropout = dials::dropout(c(0, 0.3))
        )

    grid <- grid_latin_hypercube(params, size = 8)
    folds <- vfold_cv(diamonds_train, v = 3, strata = price)

    if (!file.exists(diamonds_rds)) {
        res <- tune_grid(wf, resamples = folds, grid = grid, control = control_grid())
        saveRDS(res, diamonds_rds)
    } else {
        res <- readRDS(diamonds_rds)
    }

    if (!file.exists(diamonds_final_rds)) {
        best <- select_best(res, metric = "rmse")
        final_wf <- finalize_workflow(wf, best)
        lf <- last_fit(final_wf, split = diamonds_split, metrics = metric_set(rmse, mae, rsq))

        diamonds_final <- list(
            best = best,
            metrics = collect_metrics(lf),
            preds = collect_predictions(lf)
        )
        saveRDS(diamonds_final, diamonds_final_rds)
    }
}

if (can_run("tidymodels", "modeldata", "kerasnip", "keras3")) {
    suppressPackageStartupMessages({
        library(tidymodels)
        library(modeldata)
        library(kerasnip)
        library(keras3)
    })

    data(attrition)

    # Correct functional signatures per kerasnip docs:
    input_block <- function(input_shape) {
        layer_input(shape = input_shape, name = "features")
    }

    dense_block <- function(tensor, units = 64, dropout = 0.0) {
        tensor |>
            layer_dense(units = units, activation = "relu") |>
            layer_dropout(rate = dropout)
    }

    concat_block <- function(input_a, input_b) {
        layer_concatenate(list(input_a, input_b))
    }

    output_block <- function(tensor, num_classes) {
        tensor |> layer_dense(units = num_classes, activation = "softmax")
    }

    create_keras_functional_spec(
        model_name = "attrition_towers",
        layer_blocks = list(
            main_input = input_block,
            tower_a = inp_spec(dense_block, "main_input"),
            tower_b = inp_spec(dense_block, "main_input"),
            joined = inp_spec(concat_block, c(input_a = "tower_a", input_b = "tower_b")),
            output = inp_spec(output_block, "joined")
        ),
        mode = "classification"
    )

    split <- initial_split(attrition, strata = Attrition)
    train <- training(split)

    rec <- recipe(Attrition ~ ., data = train) |>
        step_zv(all_predictors()) |>
        step_dummy(all_nominal_predictors()) |>
        step_normalize(all_numeric_predictors())

    spec <- attrition_towers(
        tower_a_units = tune(),
        tower_b_units = tune(),
        fit_epochs = 5
    ) |>
        set_engine("keras")

    wf <- workflow() |>
        add_recipe(rec) |>
        add_model(spec)

    params <- extract_parameter_set_dials(wf) |>
        update(
            tower_a_units = dials::hidden_units(c(16, 128)),
            tower_b_units = dials::hidden_units(c(16, 128))
        )

    grid <- grid_latin_hypercube(params, size = 8)
    folds <- vfold_cv(train, v = 3, strata = Attrition)

    if (!file.exists(attrition_rds)) {
        res <- tune_grid(wf, resamples = folds, grid = grid, metrics = metric_set(roc_auc, accuracy))
        saveRDS(res, attrition_rds)
    } else {
        res <- readRDS(attrition_rds)
    }

    if (!file.exists(attrition_final_rds)) {
        best <- select_best(res, metric = "roc_auc")
        final_wf <- finalize_workflow(wf, best)
        lf <- last_fit(final_wf, split = split, metrics = metric_set(roc_auc, accuracy))

        preds <- collect_predictions(lf)
        event_level <- levels(preds$Attrition)[1]
        preferred <- paste0(".pred_", event_level)
        pos_col <- if (preferred %in% names(preds)) {
            preferred
        } else {
            pcols <- grep("^\\.pred_", names(preds), value = TRUE)
            pcols <- setdiff(pcols, ".pred_class")
            pcols[1]
        }

        attrition_final <- list(
            best = best,
            metrics = collect_metrics(lf),
            preds = preds,
            roc = if (!is.na(pos_col)) {
                roc_curve(preds, truth = Attrition, !!rlang::sym(pos_col))
            } else {
                NULL
            }
        )
        saveRDS(attrition_final, attrition_final_rds)
    }
}
```

::: {.content-visible when-profile="english"}

## kerasnip: A Bridge Between 'keras' and 'tidymodels' {.title-slide background-color="#f0f1f2"}

### Madrid R Group | February 2026
**kerasnip v0.1.0.900**

Docs: [davidrsch.github.io/kerasnip](https://davidrsch.github.io/kerasnip/) · Repo: [github.com/davidrsch/kerasnip](https://github.com/davidrsch/kerasnip)

::: {.notes}
Open with energy. Ask: "Who loves tidymodels?" then "Who fights with Keras architectures?"

Say:
- Keras and tidymodels are like oil and water — kerasnip is the emulsifier.
- Today: solving the *architectural tuning* problem.

If stuck: "Keras builds freely; tidymodels tunes systematically. We bridge them."
:::

---

## Today

- Why tuning Keras inside tidymodels is hard
- What `kerasnip` generates (and why it matters)
- Sequential workflow: structural tuning with `num_{block}`
- Functional workflow: branch/merge graphs with `inp_spec()`
- Debugging + practical tips

::: {.notes}
Goal: orient people so they don’t get lost later.

Say:
- Timing is flexible; clarity first.
- First half: sequential workflow + structural tuning (`num_{block}`) + debugging.
- Second half: functional graphs via `inp_spec()` with the same tuning workflow.

Also mention (fast):
- Debug helpers: `extract_valid_grid()`, `inform_errors()`.
- Keras object helpers: `keras_optimizers`, `keras_losses`, `keras_metrics`, `optimizer_function()`, `loss_function_keras()`.
- Spec lifecycle: `model_exists()`, `remove_keras_spec()`, `get_model_env()`.

Transition: Let’s situate Keras vs tidymodels and name the gap.
:::

---

## The Landscape

::: {.columns}
::: {.column width="50%"}
### 🧱 Keras (via `keras3`)
- **Deep Learning**
- Tensor-centric & Imperative
- Backend agnostic (TF, JAX, Torch)
- Arbitrary graphs of layers
:::

::: {.column width="50%"}
### 🧹 Tidymodels
- **Machine Learning**
- Dataframe-centric & Declarative
- Unified interface (`fit`, `predict`)
- Standardized tuning
:::
:::

**The Gap**: Using `tune_grid()` to optimize Neural Network topology (depth/width) is difficult because `parsnip` expects fixed arguments, while Keras models are arbitrary graphs.

::: {.notes}
Analogy: Keras = architect's blank sheet; tidymodels = government form with fixed fields.

Say:
- kerasnip turns the building's *structure* into form fields.
- You define the blueprint (blocks), it exposes the knobs (depth, width) as standard arguments.

If stuck: "Keras draws freely; tidymodels fills forms. kerasnip makes the forms fit the drawings."
:::

---

## The Pain Point (in one sentence)

Keras models are **arbitrary graphs**, but `parsnip` specs are **fixed argument sets**.

So you can tune learning rate easily, but tuning *model structure* (depth / width / branches) is awkward.

::: {.notes}
Goal: make the problem feel concrete and familiar.

Say:
- This is not “deep learning 101”. It’s: how to keep Keras *inside* the tidymodels workflow.
- The friction is that architecture changes live in Keras code, but tidymodels tuning expects named arguments.

Transition: So the question becomes: how do we turn topology decisions into tunable spec args?
:::

---

## Ecosystem Context

| Package | Role | Limitation |
|---------|------|------------|
| `parsnip::mlp()` | Tidymodels MLP | Mostly fixed topology; little architecture/feature engineering |
| `reticulate` | Python Bridge | Low-level, no ML workflow |
| `keras3` | R Interface | No native `tune()` integration |
| `luz`/`brulee` | Torch + Tidymodels | Torch-only backend |
| **`kerasnip`** | **Keras + Tidymodels** | **Bridges Keras flexibility with Tidymodels workflows** |

::: {.notes}
Goal: position kerasnip precisely (avoid confusion).

Say:
- Not replacing `keras3`.
- Not a new engine.
- It generates *new parsnip model specs* that still use the existing `keras` engine.

Also say explicitly:
- `parsnip::mlp()` is great for standard MLPs, but limited for architecture search.
- kerasnip focuses on tuning *structure* (depth/width/branches) without leaving tidymodels.

Transition: Now: what “dynamic specs” means in practice.
:::

---

## The Solution: Dynamic Specs

`kerasnip` is a **Model Builder**, not a Model Zoo.

1.  **Define Blocks**: Functions returning Keras layers.
2.  **Meta-program Spec**: `create_keras_*_spec()` generates a new `parsnip` model.
3.  **Tune Structure**: Block arguments (and repetition count) can be tuned.

::: {.notes}
Goal: explain the “dynamic spec” idea in one breath.

Say:
- This is a code-generation approach: you define blocks, then `kerasnip` generates a parsnip spec function.
- Once it’s a parsnip spec, the rest is standard: recipes → workflows → `tune_grid()`.

Transition: Keep the mental model in mind; we’ll come back to it on every example.
:::

---

## Mental Model {.mermaid-wide}

<br/>
<br/>
<br/>

```{mermaid}
%%| fig-align: center
 
flowchart LR
  classDef step fill:#f5f7fa,stroke:#5c6b7a,color:#1b3a57,stroke-width:2px;
  classDef tune fill:#BF281B,stroke:#BF281B,color:#ffffff,stroke-width:2px;

  A["Blocks<br/>(layer blocks)"] --> B["Spec function<br/>create_keras_*_spec()"]
  B --> C["Workflow<br/>recipe() + workflow()"]
  C --> D["Tuning<br/>tune_grid()"]
  A -. "args → tunables<br/>num_{block} repeats" .-> B

  class A,B,C step;
  class D tune;
```

::: {.notes}
Goal: anchor the whole talk to one repeatable mental model.

Say:
- If someone gets lost: blocks → spec → workflow → tuning.

Under the hood (if asked):
- Specs are registered in an internal environment.
- That’s why there are lifecycle helpers: `model_exists()`, `get_model_env()`, `remove_keras_spec()`.

Transition: Next: the minimal set of functions you’ll see in the examples.
:::

---

## Key Functions You’ll See

- `create_keras_sequential_spec()`
- `create_keras_functional_spec()`
- `inp_spec()` (functional wiring)
- `compile_keras_grid()` (debug/inspect generated graphs)
- `extract_parameter_set_dials()` (what can be tuned?)

::: {.notes}
Say:
- No need to memorize — just know each function's role.
- Other helpers exist (debugging, evaluation, housekeeping) and are covered in the docs.

Transition: Next: naming conventions, so the code reads naturally.
:::

---

## Naming Conventions (auto-generated args)

- Arguments get **prefixed** by block name (e.g. `body_units`)
- Repeatable blocks expose `num_{block}` (e.g. `num_body`)
- Repetitions **share** the same hyperparameters (by design)

Example mapping:

`dense_block(model, units = 128, dropout = 0.1)` → `body_units`, `body_dropout`

::: {.notes}
Goal: remove naming confusion before we touch code.

Say:
- Block args become spec args with prefixes (e.g. `body_units`).
- Repetition is controlled by `num_{block}`.
- Repeated blocks share hyperparameters by design.

Transition: With that mapping, the sequential example will look “normal”.
:::

---

## Sequential vs Functional (rule of thumb)

- **Sequential**: one stream of features, simple stacks
- **Functional**: multi-input, branch/merge, skip connections, explicit graphs

::: {.notes}
Goal: give a decision rule people can remember.

Say:
- Use sequential for “a chain of layers”.
- Use functional when you need an explicit graph: branches/merges, skips, multi-input.

Transition: We’ll start sequential (easier), then come back to functional.
:::

---

## Common Gotchas

- Ensure the recipe outputs numeric matrices (dummy + normalize)
- Start with a small grid, confirm structure via `compile_keras_grid()`
- Keep tuning spaces reasonable (depth × width can explode)

::: {.notes}
Say:
- Recipes must output numeric matrices (dummy + normalize).
- Start small: confirm structure with `compile_keras_grid()` before tuning.
- Keep search spaces reasonable; depth x width can explode.

If stuck: "Numeric in, small grid first, check the build."

Transition: With those guardrails, let's run a simple sequential example.
:::
---

## Workflow 1: Sequential API (Regression)

**Goal**: Predict diamond prices and tune **depth + width**.

Dataset: `ggplot2::diamonds`.

::: {.notes}
Goal: set up a simple, credible demo.

Say:
- This is a tabular MLP on diamonds. Simple on purpose: the point is tuning depth/width, not feature engineering.

Transition: First we define blocks (just functions), then generate the spec.
:::

---

## Sequential Blocks (minimal)

```r
input_block <- function(model, input_shape) {
  keras_model_sequential(input_shape = input_shape)
}

dense_block <- function(model, units = 128, dropout = 0.0) {
  model |>
    layer_dense(units = units, activation = "relu") |>
    layer_dropout(rate = dropout)
}

output_block <- function(model) {
  model |> layer_dense(units = 1)
}
```

::: {.notes}
Say:
- Blocks are plain R functions — nothing magical.
- The function signature (units, dropout) defines what becomes tunable later.
- Repeated blocks (via `num_body`) share hyperparameters by design.

If stuck: "If you want to tune it, make it an argument."
:::

---

## Generate the `parsnip` spec

```r
create_keras_sequential_spec(
  model_name = "diamond_mlp",
  layer_blocks = list(
    input = input_block,
    body = dense_block,  # <--- This block will be repeatable
    output = output_block
  ),
  mode = "regression"
)
```

::: {.notes}
Goal: make the “spec generation” moment land.

Say:
- This creates a new parsnip spec function: `diamond_mlp()`.
- From this point on, it behaves like any other tidymodels model: you can set engine, add to workflows, and tune args.

Transition: Next we’ll tune topology via `num_body`.
:::

---

## Structural Tuning

We treat **topology** as a hyperparameter.

```r
spec <- diamond_mlp(
  num_body = tune(),       # <--- Depth (e.g., 1, 2, or 3 layers)
  body_units = tune(),     # <--- Width (e.g., 64 vs 128 units)
  body_dropout = tune(),
  fit_epochs = 20
) |>
  set_engine("keras")
```

The arguments `num_{block}` are automatically generated by `kerasnip` for any repeatable block.

::: {.notes}
Goal: make “depth is tunable” feel obvious.

Say:
- `num_body` allows searching topology inside a standard `tune_grid()`.

What to emphasize:
- Depth becomes a first-class parameter.
- Keep training controls mostly fixed during architecture search.

Practical tip:
- Tune structure first (depth/width), then optionally tune optimizer/training hyperparameters.

Transition: Next: put it into a normal workflow.
:::

---

## Diamonds + Tidymodels Workflow

```r
library(tidymodels)
library(kerasnip)

data(diamonds, package = "ggplot2")

diamonds_split <- initial_split(diamonds, strata = price)
diamonds_train <- training(diamonds_split)

diamonds_rec <- recipe(price ~ ., data = diamonds_train) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

wf <- workflow() |>
  add_recipe(diamonds_rec) |>
  add_model(spec)
```

::: {.notes}
Say:
- The workflow looks identical to any other tidymodels model.
- Keras complexity is hidden behind the parsnip spec.

Transition: Next: define the tuning ranges and run `tune_grid()`.
:::

---

## Tuning Grid (including depth) {.compact .code-compact}

```r
params <- extract_parameter_set_dials(wf) |>
  update(
    num_body     = dials::num_terms(c(1, 4)),
    body_units   = dials::hidden_units(c(32, 256)),
    body_dropout = dials::dropout(c(0, 0.4))
  )

grid  <- grid_latin_hypercube(params, size = 20)
folds <- vfold_cv(diamonds_train, v = 5, strata = price)
res   <- tune_grid(wf, resamples = folds, grid = grid)
```

::: {.notes}
Goal: show this is just standard tidymodels tuning.

Say:
- `extract_parameter_set_dials()` discovers tunables, then we define ranges and build a grid.
- The key is that `num_body` is now a first-class tuning parameter.

Transition: Now let’s look at CV results and pick a candidate before touching the test set.
:::

---

## Results (Diamonds): CV — Table

```{r}
#| label: diamonds-results-cv-table
#| echo: false
#| results: asis
#| message: false
#| warning: false
diamonds_rds <- file.path("results", "diamonds_res.rds")

if (file.exists(diamonds_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
        library(tidymodels)
    })

    res <- readRDS(diamonds_rds)

    best_tbl <- show_best(res, metric = "rmse", n = 5)
    best_tbl <- best_tbl |>
        dplyr::select(num_body, body_units, body_dropout, mean, std_err) |>
        dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3)))
    cat(knitr::kable(best_tbl, format = "html"), "\n\n")
} else {
    cat("No cached results found yet. Render once to compute them.\n")
}
```

::: {.notes}
Goal: pick 1–2 strong candidates (don’t read the whole table).

Say:
- These are cross-validation estimates (mean + std_err). We’re selecting on CV, not the test set.
- I usually look for a “stable region”: low mean RMSE and not wildly sensitive to depth/dropout.

Transition: Let’s look at the CV plot to see interactions.
:::

---

## Results (Diamonds): CV — Plot

```{r}
#| label: diamonds-results-cv-plot
#| echo: false
#| message: false
#| warning: false
diamonds_rds <- file.path("results", "diamonds_res.rds")

if (file.exists(diamonds_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
        library(ggplot2)
        library(tidymodels)
    })

    res <- readRDS(diamonds_rds)

    p_trends <- collect_metrics(res) |>
        filter(.metric == "rmse") |>
        ggplot(aes(x = num_body, y = mean, color = factor(body_units), group = interaction(body_units, body_dropout))) +
        geom_point() +
        geom_line(alpha = 0.35) +
        facet_wrap(~body_dropout) +
        labs(
            x = "num_body (depth)",
            y = "CV RMSE (mean)",
            color = "body_units",
            title = "Cross-validation trends across depth/width"
        )
    print(p_trends)
} else {
    cat("No cached results found yet. Render once to compute them.\n")
}
```

::: {.notes}
Goal: see interactions (depth × width × dropout) at a glance.

Say:
- If the trend flattens, deeper models aren’t buying much.
- When differences are tiny relative to noise, pick simpler.

Transition: Now we do a one-time test-set confirmation.
:::

---

## Results (Diamonds): Test — Table

```{r}
#| label: diamonds-results-test-table
#| echo: false
#| results: asis
#| message: false
#| warning: false
diamonds_final_rds <- file.path("results", "diamonds_final.rds")

if (file.exists(diamonds_final_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
    })

    final <- readRDS(diamonds_final_rds)

    metrics_tbl <- final$metrics |>
        dplyr::select(.metric, .estimate) |>
        dplyr::mutate(.estimate = round(.estimate, 4))
    cat(knitr::kable(metrics_tbl, format = "html"), "\n\n")
} else {
    cat("No cached final results found yet.\n")
}
```

::: {.notes}
Goal: one-time confirmation.

Say:
- Test metrics are for final reporting; avoid using them to decide hyperparameters.
- If test is worse than CV by a lot, it’s a sign of instability or leakage.

Transition: I’ll show a quick visual sanity check on test predictions.
:::

---

## Results (Diamonds): Test — Plot

```{r}
#| label: diamonds-results-test-plot
#| echo: false
#| message: false
#| warning: false
diamonds_final_rds <- file.path("results", "diamonds_final.rds")

if (file.exists(diamonds_final_rds)) {
    suppressPackageStartupMessages({
        library(ggplot2)
    })

    final <- readRDS(diamonds_final_rds)

    p_test <- final$preds |>
        ggplot(aes(x = price, y = .pred)) +
        geom_point(alpha = 0.25) +
        geom_abline(slope = 1, intercept = 0, linewidth = 0.8) +
        coord_fixed() +
        labs(x = "Truth (price)", y = "Predicted", title = "Final model on test set")
    print(p_test)
} else {
    cat("No cached final results found yet.\n")
}
```

::: {.notes}
Say:
- Points near the diagonal = good predictions; systematic deviations reveal bias.
- High-price diamonds are typically harder to predict — check spread there.

If stuck: "Tight diagonal = strong fit; fan-out = heteroscedasticity."

Transition: Next: how to debug/inspect the generated architecture.
:::

---

## Debugging: “What model did we just create?” {.compact .table-compact}

```r
# compile_keras_grid(spec, grid, x, y)
```

```{r}
#| label: diamonds-debug
#| echo: false
#| message: false
#| warning: false
suppressPackageStartupMessages({
    library(dplyr)
    library(tibble)
    library(purrr)
    library(tidymodels)
    library(kerasnip)
    library(keras3)
})

if (!exists("diamond_mlp", mode = "function") || !kerasnip::model_exists("diamond_mlp")) {
    input_block <- function(model, input_shape) {
        keras_model_sequential(input_shape = input_shape)
    }
    dense_block <- function(model, units = 128, dropout = 0.0) {
        model |>
            layer_dense(units = units, activation = "relu") |>
            layer_dropout(rate = dropout)
    }
    output_block <- function(model) {
        model |> layer_dense(units = 1)
    }

    create_keras_sequential_spec(
        model_name = "diamond_mlp",
        layer_blocks = list(input = input_block, body = dense_block, output = output_block),
        mode = "regression"
    )
}

prep_rec <- prep(diamonds_rec, training = diamonds_train)
juiced <- juice(prep_rec)
x <- juiced |> select(-price)
y <- juiced$price

dbg <- compile_keras_grid(
    spec = diamond_mlp(
        num_body = tune(),
        body_units = tune(),
        body_dropout = tune(),
        fit_epochs = 5
    ) |> set_engine("keras"),
    grid = tibble(num_body = 3, body_units = 128, body_dropout = 0.2),
    x = x,
    y = y
)

dbg |> select(num_body, body_units, body_dropout, error)

if (!is.null(dbg$compiled_model[[1]])) {
    model <- dbg$compiled_model[[1]]
    tibble(
        n_features = ncol(x),
        n_layers = length(model$layers),
        n_params = model$count_params()
    )

    layer_tbl <- tibble(
        layer = map_chr(model$layers, "name"),
        class = map_chr(model$layers, ~ class(.x)[1])
    ) |>
        slice_head(n = 8)
    layer_tbl
}
```

::: {.notes}
Say:
- `compile_keras_grid()` confirms depth/width changes are actually happening.
- `extract_valid_grid()` keeps only successful builds; `inform_errors()` summarizes failures.

Transition: Next: switch to a functional (graph) example.
:::
---

## Workflow 2: Functional API (Branch/Merge)

**Goal**: Employee Attrition (Classification).

Dataset: `modeldata::attrition`.

We’ll build a small **branch/merge** graph:
- one input tensor
- two parallel “towers”
- concat + output

::: {.notes}
Goal: motivate why functional Keras matters.

Say:
- Functional is the sweet spot when the architecture is not a straight line.

Important practical choice for a live talk: keep preprocessing simple.
We use ONE input produced by a standard recipe, then create two parallel branches from the same input.

Transition: `inp_spec()` is the wiring mechanism.
:::

---

## Functional Wiring with `inp_spec()`

`inp_spec(func, inputs)` explicitly maps **upstream block outputs** to **function arguments**.

That’s the key to:
- multi-input models
- skip connections
- branch/merge architectures

::: {.notes}
Analogy: patch cables on a synthesizer. `inp_spec()` labels each cable.

Say:
- In Keras you assign variables (`x = layer(input)`). In kerasnip you declare connections by name.
- `inp_spec(block, "name")` = "plug Tower A's output into Concat's input".

If stuck: "It's just labeling which output goes into which input."
:::
---

## Functional Blocks (minimal) {.compact .code-compact}

```r
# In functional Keras, tensors flow through your blocks.
input_block <- function(input_shape) {
  layer_input(shape = input_shape, name = "features")
}

dense_block <- function(tensor, units = 64, dropout = 0.0) {
  tensor |>
    layer_dense(units = units, activation = "relu") |>
    layer_dropout(rate = dropout)
}

concat_block <- function(input_a, input_b) {
  layer_concatenate(list(input_a, input_b))
}

output_block <- function(tensor, num_classes) {
  tensor |> layer_dense(units = num_classes, activation = "softmax")
}
```

::: {.notes}
Say:
- In functional Keras, blocks receive tensors (not models). Input block takes `input_shape`; downstream blocks take the upstream tensor.
- `inp_spec()` wires outputs to inputs by name; names must match exactly.

If stuck: "Functional = tensors flowing through a graph; sequential = layers stacking on a model."
:::
---

## Build the graph (spec) {.compact .code-compact}

```r
create_keras_functional_spec(
  model_name = "attrition_towers",
  layer_blocks = list(
    main_input = input_block,
    tower_a = inp_spec(dense_block, "main_input"),
    tower_b = inp_spec(dense_block, "main_input"),
    joined = inp_spec(concat_block, c(input_a = "tower_a", input_b = "tower_b")),
    output = inp_spec(output_block, "joined")
  ),
  mode = "classification"
)
```

::: {.notes}
Goal: show functional graphs still fit a standard workflow.

Say:
- This gives us a non-linear graph while still using a standard tidymodels recipe.

If your architecture is a straight line, sequential is simpler.
Functional is for “graphs”: branches, merges, skips.

If asked about multi-input:
- Define multiple `layer_input()` blocks and wire them into downstream blocks by name using `inp_spec()`.

Transition: Next: build the workflow and tune.
:::

---

## Attrition Workflow (high level)

```r
library(tidymodels)
library(modeldata)

data(attrition)

attr_split <- initial_split(attrition, strata = Attrition)
attr_train <- training(attr_split)

attr_rec <- recipe(Attrition ~ ., data = attr_train) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

attr_spec <- attrition_towers(
  tower_a_units = tune(),
  tower_b_units = tune(),
  fit_epochs    = 20
) |>
  set_engine("keras")

attr_wf <- workflow() |>
  add_recipe(attr_rec) |>
  add_model(attr_spec)
```

::: {.notes}
Say:
- Same workflow pattern: recipe + spec + workflow + tune_grid().
- Ensure the outcome is a factor and levels are stable.

Transition: Now: CV results and selection.
:::
---

## Results (Attrition): CV — Table

::: {.notes}
Say:
- These are cross-validation estimates (mean + std_err). Select on CV, not on test.
- Look for a stable region: good AUC with low sensitivity to tower sizes.

Transition: Let's look at the CV plot to see interactions.
:::
```{r}
#| label: attrition-results-cv-table
#| echo: false
#| results: asis
#| message: false
#| warning: false
attrition_rds <- file.path("results", "attrition_res.rds")

if (file.exists(attrition_rds)) {
    attr_res <- readRDS(attrition_rds)
    suppressPackageStartupMessages({
        library(dplyr)
        library(tune)
    })

    best_tbl <- show_best(attr_res, metric = "roc_auc", n = 5)
    best_tbl <- best_tbl |>
        dplyr::select(tower_a_units, tower_b_units, mean, std_err) |>
        dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 4)))
    cat(knitr::kable(best_tbl, format = "html"), "\n\n")
} else {
    cat("No cached results found yet. Render once to compute them.\n")
}
```

::: {.notes}
Say:
- Look at mean ROC AUC but also std_err.
- Flat heatmap = robust architecture; pick the cheaper/smaller model.
- Sharp peak = sensitive; might be a local optimum.
- Rule of thumb: pick the simplest within 1 std_err of the best.
:::
---

## Results (Attrition): CV — Plot

```{r}
#| label: attrition-results-cv-plot
#| echo: false
#| message: false
#| warning: false
attrition_rds <- file.path("results", "attrition_res.rds")

if (file.exists(attrition_rds)) {
    attr_res <- readRDS(attrition_rds)
    suppressPackageStartupMessages({
        library(dplyr)
        library(ggplot2)
        library(tune)
    })

    p_heat <- collect_metrics(attr_res) |>
        filter(.metric == "roc_auc") |>
        ggplot(aes(x = factor(tower_a_units), y = factor(tower_b_units), fill = mean)) +
        geom_tile() +
        scale_fill_gradient(low = "#f0f1f2", high = "#BF281B") +
        labs(x = "tower_a_units", y = "tower_b_units", fill = "mean roc_auc", title = "CV mean roc_auc")
    print(p_heat)
} else {
    cat("No cached results found yet. Render once to compute them.\n")
}
```

::: {.notes}
Goal: see whether one tower matters more than the other.

Say:
- A flat heatmap means tuning isn’t very sensitive.
- A sharp peak means you should narrow the search around that region.

Transition: Now: one-time test reporting.
:::

---

## Results (Attrition): Test — Table

```{r}
#| label: attrition-results-test-table
#| echo: false
#| results: asis
#| message: false
#| warning: false
attrition_final_rds <- file.path("results", "attrition_final.rds")

if (file.exists(attrition_final_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
    })

    final <- readRDS(attrition_final_rds)

    metrics_tbl <- final$metrics |>
        dplyr::select(.metric, .estimate) |>
        dplyr::mutate(.estimate = round(.estimate, 4))
    cat(knitr::kable(metrics_tbl, format = "html"), "\n\n")
} else {
    cat("No cached final results found yet.\n")
}
```

::: {.notes}
Goal: report final test performance once.

Say:
- CV chose the architecture; test is for final confirmation.
- If the dataset is imbalanced, ROC AUC is often more informative than accuracy.

Transition: Show ROC curve as a visual summary.
:::

---

## Results (Attrition): Test — Plot

```{r}
#| label: attrition-results-test-plot
#| echo: false
#| message: false
#| warning: false
attrition_final_rds <- file.path("results", "attrition_final.rds")

if (file.exists(attrition_final_rds)) {
    suppressPackageStartupMessages({
        library(ggplot2)
    })

    final <- readRDS(attrition_final_rds)

    if (!is.null(final$roc)) {
        p_roc <- ggplot(final$roc, aes(x = 1 - specificity, y = sensitivity)) +
            geom_line(linewidth = 1) +
            geom_abline(linetype = "dashed", alpha = 0.5) +
            coord_equal() +
            labs(x = "False positive rate", y = "True positive rate", title = "ROC curve (test set)")
        print(p_roc)
    } else {
        cat("ROC curve not available.\n")
    }
} else {
    cat("No cached final results found yet.\n")
}
```

::: {.notes}
Say:
- The further the curve from the diagonal, the better the discrimination.
- When results are close, pick the simplest architecture within uncertainty.

Transition: Close with practical tips and failure modes.
:::
---

## Practical Tips (so you don’t suffer)

- Start with `compile_keras_grid()` before tuning
- Keep blocks small and testable
- Prefer a few meaningful tunables over 30 tiny ones

::: {.notes}
Say:
- If something breaks in a functional graph, reduce to 2-3 blocks to isolate the error.
- Keep blocks small and testable; prefer a few meaningful tunables over many tiny ones.

Transition: I'll close with trade-offs and the roadmap.
:::
---

## Advantages & Limitations

### Advantages
1.  **Workflow Parity**: Use `recipes`, `workflows`, and `tune` exactly like `glmnet`.
2.  **Reproducibility**: Topology is code. Version control your architecture.
3.  **Extensibility**: Wrap *any* Keras layer (Transformers, RNNs) into a block.

### Limitations
-   **Manual Blocking**: No pre-shipped "ResNet50" blocks yet (you must define them).
-   **Custom Losses**: Integration requires some boilerplate for custom objects.

::: {.notes}
Say:
- The learning curve is writing good blocks; after that, the workflow is standard.
- No pre-built blocks yet (e.g. ResNet50), but any Keras layer can be wrapped.

Transition: Close with a short roadmap and invite questions.
:::
---

## Roadmap

- 📦 **Templates**: High-level helpers for common architectures.
- 🔧 **Fine-tuning**: Workflows for adapting pre-trained models.

### Resources
- 📦 **Package**: [github.com/davidrsch/kerasnip](https://github.com/davidrsch/kerasnip)
- 📑 **Docs**: [davidrsch.github.io/kerasnip](https://davidrsch.github.io/kerasnip/)

::: {.notes}
Goal: land the talk and create a clean Q&A opening.

Say:
- Recap: blocks → spec → workflow → tuning.
- Invite feedback and contributions.
- Leave time for Q&A.

Transition: Stop sharing, take questions.
:::

:::

::: {.content-visible when-profile="spanish"}

## kerasnip: Un Puente entre 'keras' y 'tidymodels' {.title-slide background-color="#f0f1f2"}

### Grupo de Usuarios de R de Madrid | Febrero 2026
**kerasnip v0.1.0.900**

Docs: [davidrsch.github.io/kerasnip](https://davidrsch.github.io/kerasnip/) · Repo: [github.com/davidrsch/kerasnip](https://github.com/davidrsch/kerasnip)

::: {.notes}
Empezar con energia. Preguntar: "Quien ama tidymodels?" y "Quien pelea con Keras?"

Decir:
- Keras y tidymodels son como aceite y agua; kerasnip es el emulsionante.
- Hoy: resolver el problema del *ajuste de arquitectura*.

Si te bloqueas: "Keras construye libre; tidymodels ajusta sistematicamente. Los conectamos."
:::
---

## Hoy

- Por qué es difícil ajustar modelos Keras dentro de tidymodels
- Qué genera `kerasnip` (y por qué importa)
- Flujo secuencial: ajuste estructural con `num_{block}`
- Flujo funcional: grafos con ramas y uniones con `inp_spec()`
- Depuración y consejos prácticos

::: {.notes}
Decir:
- Primera parte: secuencial + ajuste estructural + depuracion.
- Segunda parte: grafos funcionales con `inp_spec()`.

Transicion: Situamos Keras y tidymodels y nombramos la brecha.
:::
---

## El Panorama

::: {.columns}
::: {.column width="50%"}
### 🧱 Keras (vía `keras3`)
- **Aprendizaje profundo**
- Pensado para tensores
- Agnóstico al backend (TF, JAX, Torch)
- Grafos arbitrarios de capas
:::

::: {.column width="50%"}
### 🧹 tidymodels
- **Aprendizaje automático**
- Pensado para data frames
- Interfaz unificada (`fit`, `predict`)
- Ajuste de hiperparámetros, de forma estandarizada
:::
:::

**La brecha**: optimizar la topología (profundidad/ancho) con `tune_grid()` es difícil: `parsnip` espera un conjunto fijo de argumentos, mientras que Keras permite grafos arbitrarios.

::: {.notes}
Analogia: Keras = mesa de arquitecto (papel en blanco); tidymodels = formulario con casillas fijas.

Decir:
- kerasnip convierte la *estructura del edificio* en casillas del formulario.
- Defines los planos (bloques), el paquete expone los diales (profundidad, ancho).

Si te bloqueas: "Keras dibuja libre; tidymodels rellena formularios. kerasnip adapta los formularios a los dibujos."
:::
---

## El Problema (en una frase)

Los modelos Keras son **grafos arbitrarios**, pero los specs de `parsnip` son **conjuntos fijos de argumentos**.

El learning rate se ajusta fácilmente, pero ajustar la *estructura* (profundidad/ancho/ramas) suele ser incómodo.

::: {.notes}
Objetivo: que el problema se entienda en 10 segundos.

Qué decir:
- Esta charla no es “deep learning 101”. Es: cómo mantener Keras *dentro* del flujo de tidymodels.
- La fricción está en que la arquitectura vive en código Keras, pero el ajuste en tidymodels espera argumentos con nombre.

Transición: Entonces: ¿cómo convertimos decisiones de topología en argumentos ajustables?
:::

---

## Contexto del Ecosistema

| Paquete | Rol | Limitación |
|---------|-----|------------|
| `parsnip::mlp()` | MLP en tidymodels | Topología casi fija; poca flexibilidad para arquitectura o ingeniería de variables |
| `reticulate` | Puente Python | Bajo nivel; sin integración con workflows/tune |
| `keras3` | Interfaz R | Sin integración nativa `tune()` |
| `luz`/`brulee` | Torch + tidymodels | Solo backend Torch |
| **`kerasnip`** | **Keras + tidymodels** | **Integra flexibilidad de Keras con workflows de tidymodels** |

::: {.notes}
Objetivo: posicionar kerasnip con precisión (sin malentendidos).

Qué decir:
- No sustituye a `keras3`.
- No es un engine nuevo.
- Genera *nuevos specs de parsnip* que usan el engine `keras` existente.

Y conviene decirlo claro:
- `parsnip::mlp()` va genial para MLPs “estándar”, pero es limitado para búsqueda arquitectónica.
- kerasnip está pensado para ajustar *estructura* (profundidad/ancho/ramas) sin salir de tidymodels.

Transición: Ahora: qué significa “spec dinámico” en la práctica.
:::

---

## La Solución: Specs Dinámicos

`kerasnip` es un **constructor de modelos**, no un “model zoo”.

1.  **Definir bloques**: funciones que devuelven capas Keras.
2.  **Generar el spec**: `create_keras_*_spec()` crea un modelo de `parsnip`.
3.  **Ajustar la estructura**: los argumentos (y el número de repeticiones) pasan a ser hiperparámetros.

::: {.notes}
Objetivo: explicar “specs dinámicos” sin jerga.

Qué decir:
- Es un enfoque de generación de código: defines bloques y `kerasnip` genera un spec de parsnip.
- A partir de ahí, todo vuelve a ser estándar: recipes → workflows → `tune_grid()`.

Transición: Mantén el modelo mental en mente: lo reutilizaremos en los ejemplos.
:::

---

## Modelo Mental {.mermaid-wide}

<br/>
<br/>
<br/>

```{mermaid}
%%| fig-align: center

flowchart LR
  classDef step fill:#f5f7fa,stroke:#5c6b7a,color:#1b3a57,stroke-width:2px;
  classDef tune fill:#BF281B,stroke:#BF281B,color:#ffffff,stroke-width:2px;

  A["Bloques<br/>(layer blocks)"] --> B["Función spec<br/>create_keras_*_spec()"]
  B --> C["Workflow<br/>recipe() + workflow()"]
  C --> D["Ajuste<br/>tune_grid()"]
  A -. "args → ajustables<br/>num_{block} repite" .-> B

  class A,B,C step;
  class D tune;
```

::: {.notes}
Objetivo: anclar toda la charla a un modelo mental repetible.

Qué decir:
- Si alguien se pierde: bloques → spec → workflow → ajuste.

Por dentro (si sale en preguntas):
- Los specs se registran en un entorno interno.
- Por eso existen helpers de ciclo de vida: `model_exists()`, `get_model_env()`, `remove_keras_spec()`.

Transición: Siguiente: funciones mínimas que verás en los ejemplos.
:::

---

## Funciones Clave

- `create_keras_sequential_spec()`
- `create_keras_functional_spec()`
- `inp_spec()` (conexiones funcionales)
- `compile_keras_grid()` (inspección/depuración)
- `extract_parameter_set_dials()` (¿qué se puede ajustar?)

::: {.notes}
Decir:
- No hay que memorizar; solo entender el rol de cada funcion.
- Existen otros helpers (depuracion, evaluacion, limpieza) documentados en la web.

Transicion: Siguiente: convenciones de nombres.
:::
---

## Convenciones de nombres (args autogenerados)

- Los argumentos se **prefijan** con el nombre del bloque (p. ej., `body_units`)
- Los bloques repetibles exponen `num_{block}` (p. ej., `num_body`)
- Las repeticiones **comparten** hiperparámetros (a propósito)

Ejemplo de mapeo:

`dense_block(model, units = 128, dropout = 0.1)` → `body_units`, `body_dropout`

::: {.notes}
Objetivo: eliminar confusión de nombres antes del código.

Qué decir:
- Los args del bloque pasan a ser args del spec con prefijo (p. ej. `body_units`).
- La repetición se controla con `num_{block}`.
- Las repeticiones comparten hiperparámetros a propósito.

Transición: Con ese mapeo, el ejemplo secuencial se ve “normal”.
:::

---

## Secuencial vs Funcional (regla práctica)

- **Secuencial**: un flujo de variables, apilados sencillos
- **Funcional**: multi-entrada, ramas/uniones, conexiones de salto (skip connections), grafos explícitos

::: {.notes}
Objetivo: dar una regla práctica fácil de recordar.

Qué decir:
- Secuencial para “una cadena de capas”.
- Funcional cuando necesitas un grafo explícito: ramas/uniones, skip connections, multi-entrada.

Transición: Empezamos con secuencial (más simple) y luego volvemos a funcional.
:::

---

## Errores Comunes

- Asegúrate de que la receta produzca matrices numéricas (dummy + normalize)
- Empieza con un grid pequeño y confirma la estructura con `compile_keras_grid()`
- Mantén el espacio de ajuste razonable (profundidad × ancho explota)

::: {.notes}
Decir:
- Las recetas deben producir matrices numericas (dummy + normalize).
- Empieza con un grid pequeno y confirma la estructura con `compile_keras_grid()`.
- Manten el espacio de busqueda razonable; profundidad x ancho explota.

Si te bloqueas: "Numerico, grid pequeno primero, revisa el build."

Transicion: Con estos guardarrailes, vamos al ejemplo secuencial.
:::
---

## Flujo 1: API Secuencial (Regresión)

**Objetivo**: Predecir precios de diamantes y ajustar **profundidad + ancho**.

Dataset: `ggplot2::diamonds`.

::: {.notes}
Objetivo: un demo simple pero creíble.

Qué decir:
- MLP tabular con diamonds. Simple a propósito: el foco es ajustar profundidad/ancho, no exprimir el feature engineering.

Transición: Primero definimos bloques (funciones), luego generamos el spec.
:::

---

## Bloques Secuenciales (mínimos)

```r
input_block <- function(model, input_shape) {
  keras_model_sequential(input_shape = input_shape)
}

dense_block <- function(model, units = 128, dropout = 0.0) {
  model |>
    layer_dense(units = units, activation = "relu") |>
    layer_dropout(rate = dropout)
}

output_block <- function(model) {
  model |> layer_dense(units = 1)
}
```

::: {.notes}
Decir:
- Los bloques son funciones normales de R; la firma define que sera ajustable.
- Las repeticiones (via `num_body`) comparten hiperparametros a proposito.

Si te bloqueas: "Si quieres ajustarlo, hazlo argumento de la funcion."

Transicion: Siguiente: generar el spec a partir de estos bloques.
:::
---

## Generar el spec `parsnip`

```r
create_keras_sequential_spec(
  model_name = "diamond_mlp",
  layer_blocks = list(
    input = input_block,
    body = dense_block,  # <--- Bloque repetible
    output = output_block
  ),
  mode = "regression"
)
```

::: {.notes}
Objetivo: que el “momento generación de spec” se entienda.

Qué decir:
- Esto crea una nueva función spec de parsnip: `diamond_mlp()`.
- A partir de aquí se comporta como cualquier modelo de tidymodels: engine, workflow y tuning.

Transición: Siguiente: ajustar topología con `num_body`.
:::

---

## Ajuste Estructural

Tratamos la **topología** como hiperparámetro.

```r
spec <- diamond_mlp(
  num_body = tune(),       # <--- Profundidad (ej. 1, 2, o 3 capas)
  body_units = tune(),     # <--- Ancho (ej. 64 vs 128 unidades)
  body_dropout = tune(),
  fit_epochs = 20
) |>
  set_engine("keras")
```

Los argumentos `num_{block}` son generados automáticamente por `kerasnip` para cualquier bloque repetible.

::: {.notes}
Objetivo: que `num_body` se entienda como hiperparámetro de topología.

Qué decir:
- Valor clave: `num_body` permite buscar topología dentro de un `tune_grid()` estándar.
- La profundidad pasa a ser un hiperparámetro más.
- Durante la búsqueda arquitectónica, conviene mantener el entrenamiento bastante fijo.
- Primero estructura (profundidad/ancho), luego (si hace falta) hiperparámetros de entrenamiento.

Transición: Siguiente: meterlo en un workflow normal.
:::

---

## Diamonds + Workflow Tidymodels

```r
library(tidymodels)
library(kerasnip)

data(diamonds, package = "ggplot2")

diamonds_split <- initial_split(diamonds, strata = price)
diamonds_train <- training(diamonds_split)

diamonds_rec <- recipe(price ~ ., data = diamonds_train) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

wf <- workflow() |>
  add_recipe(diamonds_rec) |>
  add_model(spec)
```

::: {.notes}
Decir:
- El workflow se ve identico a cualquier otro modelo de tidymodels.
- La complejidad Keras queda detras del spec parsnip.

Transicion: Siguiente: definir rangos y lanzar `tune_grid()`.
:::
---

## Grid de ajuste (incluye profundidad) {.compact .code-compact}

```r
params <- extract_parameter_set_dials(wf) |>
  update(
    num_body     = dials::num_terms(c(1, 4)),
    body_units   = dials::hidden_units(c(32, 256)),
    body_dropout = dials::dropout(c(0, 0.4))
  )

grid  <- grid_latin_hypercube(params, size = 20)
folds <- vfold_cv(diamonds_train, v = 5, strata = price)
res   <- tune_grid(wf, resamples = folds, grid = grid)
```

::: {.notes}
Objetivo: mostrar que el ajuste es tidymodels estándar.

Qué decir:
- `extract_parameter_set_dials()` detecta qué se puede ajustar; luego definimos rangos y construimos un grid.
- Lo importante: `num_body` ya es un hiperparámetro de primera clase.

Transición: Ahora vemos resultados en CV y elegimos candidato antes de mirar el test.
:::

---

## Resultados (Diamantes): CV — Tabla

```{r}
#| label: diamonds-results-cv-table-es
#| echo: false
#| results: asis
#| message: false
#| warning: false
diamonds_rds <- file.path("results", "diamonds_res.rds")

if (file.exists(diamonds_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
        library(tidymodels)
    })

    res <- readRDS(diamonds_rds)

    best_tbl <- show_best(res, metric = "rmse", n = 5)
    best_tbl <- best_tbl |>
        dplyr::select(num_body, body_units, body_dropout, mean, std_err) |>
        dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3)))
    cat(knitr::kable(best_tbl, format = "html"), "\n\n")
} else {
    cat("Todavía no hay resultados en caché. Renderiza una vez para calcularlos.\n")
}
```

::: {.notes}
Objetivo: elegir 1–2 candidatos (no leer toda la tabla).

Qué decir:
- Son estimaciones de validación cruzada (media + error estándar). Se selecciona con CV, no con el test.
- Busca una zona “estable”: buen RMSE medio y poca sensibilidad a profundidad/dropout.
Transición: Veamos el gráfico para entender interacciones.
:::

---

## Resultados (Diamantes): CV — Gráfico

```{r}
#| label: diamonds-results-cv-plot-es
#| echo: false
#| message: false
#| warning: false
diamonds_rds <- file.path("results", "diamonds_res.rds")

if (file.exists(diamonds_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
        library(ggplot2)
        library(tidymodels)
    })

    res <- readRDS(diamonds_rds)

    p_trends <- collect_metrics(res) |>
        filter(.metric == "rmse") |>
        ggplot(aes(x = num_body, y = mean, color = factor(body_units), group = interaction(body_units, body_dropout))) +
        geom_point() +
        geom_line(alpha = 0.35) +
        facet_wrap(~body_dropout) +
        labs(
            x = "num_body (profundidad)",
            y = "RMSE CV (media)",
            color = "body_units",
            title = "Tendencias en validación cruzada"
        )
    print(p_trends)
} else {
    cat("Todavía no hay resultados en caché. Renderiza una vez para calcularlos.\n")
}
```

::: {.notes}
Objetivo: ver interacciones (profundidad × ancho × dropout) de un vistazo.

Qué decir:
- Si la tendencia se aplana, más profundidad no aporta.
- Si las diferencias son pequeñas, elige lo más simple.
Transición: Ahora confirmamos una sola vez en el conjunto de test.
:::

---

## Resultados (Diamantes): Test — Tabla

```{r}
#| label: diamonds-results-test-table-es
#| echo: false
#| results: asis
#| message: false
#| warning: false
diamonds_final_rds <- file.path("results", "diamonds_final.rds")

if (file.exists(diamonds_final_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
    })

    final <- readRDS(diamonds_final_rds)

    metrics_tbl <- final$metrics |>
        dplyr::select(.metric, .estimate) |>
        dplyr::mutate(.estimate = round(.estimate, 4))
    cat(knitr::kable(metrics_tbl, format = "html"), "\n\n")
} else {
    cat("Todavía no hay resultados finales en caché.\n")
}
```

::: {.notes}
Objetivo: confirmación final una sola vez.

Qué decir:
- El test es para reportar, no para decidir hiperparámetros.
- Si el test empeora mucho respecto a CV, sospecha inestabilidad o leakage.
Transición: Siguiente: una comprobación visual rápida en las predicciones de test.
:::

---

## Resultados (Diamantes): Test — Gráfico

```{r}
#| label: diamonds-results-test-plot-es
#| echo: false
#| message: false
#| warning: false
diamonds_final_rds <- file.path("results", "diamonds_final.rds")

if (file.exists(diamonds_final_rds)) {
    suppressPackageStartupMessages({
        library(ggplot2)
    })

    final <- readRDS(diamonds_final_rds)

    p_test <- final$preds |>
        ggplot(aes(x = price, y = .pred)) +
        geom_point(alpha = 0.25) +
        geom_abline(slope = 1, intercept = 0, linewidth = 0.8) +
        coord_fixed() +
        labs(x = "Valor real (price)", y = "Predicho", title = "Modelo final en test")
    print(p_test)
} else {
    cat("Todavía no hay resultados finales en caché.\n")
}
```

::: {.notes}
Decir:
- Puntos cerca de la diagonal = buenas predicciones; desviaciones sistematicas revelan sesgo.
- Los diamantes caros suelen ser mas dificiles de predecir; revisa la dispersion ahi.

Si te bloqueas: "Diagonal ajustada = buen modelo; abanico = heterocedasticidad."

Transicion: Siguiente: como depurar/inspeccionar la arquitectura generada.
:::
---

## Depuración: “¿qué modelo se generó?” {.compact .table-compact}

```r
# compile_keras_grid(spec, grid, x, y)
```

```{r}
#| label: diamonds-debug-es
#| echo: false
#| message: false
#| warning: false
suppressPackageStartupMessages({
    library(dplyr)
    library(tibble)
    library(purrr)
    library(tidymodels)
    library(kerasnip)
    library(keras3)
})

if (!exists("diamond_mlp", mode = "function") || !kerasnip::model_exists("diamond_mlp")) {
    input_block <- function(model, input_shape) {
        keras_model_sequential(input_shape = input_shape)
    }
    dense_block <- function(model, units = 128, dropout = 0.0) {
        model |>
            layer_dense(units = units, activation = "relu") |>
            layer_dropout(rate = dropout)
    }
    output_block <- function(model) {
        model |> layer_dense(units = 1)
    }

    create_keras_sequential_spec(
        model_name = "diamond_mlp",
        layer_blocks = list(input = input_block, body = dense_block, output = output_block),
        mode = "regression"
    )
}

prep_rec <- prep(diamonds_rec, training = diamonds_train)
juiced <- juice(prep_rec)
x <- juiced |> select(-price)
y <- juiced$price

dbg <- compile_keras_grid(
    spec = diamond_mlp(
        num_body = tune(),
        body_units = tune(),
        body_dropout = tune(),
        fit_epochs = 5
    ) |> set_engine("keras"),
    grid = tibble(num_body = 3, body_units = 128, body_dropout = 0.2),
    x = x,
    y = y
)

dbg |> select(num_body, body_units, body_dropout, error)

if (!is.null(dbg$compiled_model[[1]])) {
    model <- dbg$compiled_model[[1]]
    tibble(
        n_features = ncol(x),
        n_layers = length(model$layers),
        n_params = model$count_params()
    )

    layer_tbl <- tibble(
        layer = map_chr(model$layers, "name"),
        class = map_chr(model$layers, ~ class(.x)[1])
    ) |>
        slice_head(n = 8)
    layer_tbl
}
```

::: {.notes}
Decir:
- `compile_keras_grid()` confirma que los cambios de estructura realmente ocurren.
- `extract_valid_grid()` filtra builds correctos; `inform_errors()` resume errores comunes.

Transicion: Siguiente: pasamos a un ejemplo funcional (grafo).
:::
---

## Flujo 2: API Funcional (Ramificar/Unir)

**Meta**: Predecir rotación de personal (attrition) — clasificación.

Dataset: `modeldata::attrition`.

Construimos un grafo sencillo de **ramas y unión**:
- una entrada
- dos “torres” en paralelo
- concat + salida

::: {.notes}
Objetivo: motivar por qué el API funcional importa.

Qué decir:
- El valor del API funcional es que dejas de estar limitado a una cadena lineal.
- Decisión práctica para que sea reproducible: usamos UNA entrada desde una receta estándar y creamos dos ramas desde el mismo tensor.

Transición: `inp_spec()` es el mecanismo de cableado.
:::

---

## Conexiones con `inp_spec()`

`inp_spec(func, inputs)` mapea explícitamente **salidas de bloques** a **argumentos**.

Útil para:
- modelos multi-entrada
- skip connections
- arquitecturas con ramas y uniones

::: {.notes}
Analogia: cables de sintetizador modular. `inp_spec()` etiqueta cada cable.

Decir:
- En Keras asignas variables (`x = capa(entrada)`). En kerasnip declaras conexiones por nombre.
- `inp_spec(bloque, "nombre")` = "enchufa la salida de Torre A en la entrada de Concat".

Si te bloqueas: "Solo es etiquetar que salida va a que entrada."
:::
---

## Bloques Funcionales (mínimos) {.compact .code-compact}

```r
# En Keras funcional, los tensores van “pasando” por bloques.
input_block <- function(input_shape) {
  layer_input(shape = input_shape, name = "features")
}

dense_block <- function(tensor, units = 64, dropout = 0.0) {
  tensor |>
    layer_dense(units = units, activation = "relu") |>
    layer_dropout(rate = dropout)
}

concat_block <- function(input_a, input_b) {
  layer_concatenate(list(input_a, input_b))
}

output_block <- function(tensor, num_classes) {
  tensor |> layer_dense(units = num_classes, activation = "softmax")
}
```

::: {.notes}
Decir:
- En Keras funcional, los bloques reciben tensores (no modelos). El bloque de entrada recibe `input_shape`; los demas reciben el tensor upstream.
- `inp_spec()` conecta salidas a entradas por nombre; los nombres deben coincidir exactamente.

Si te bloqueas: "Funcional = tensores fluyendo por un grafo; secuencial = capas apiladas sobre un modelo."
:::
---

## Construir el grafo (spec) {.compact .code-compact}

```r
create_keras_functional_spec(
  model_name = "attrition_towers",
  layer_blocks = list(
    main_input = input_block,
    tower_a = inp_spec(dense_block, "main_input"),
    tower_b = inp_spec(dense_block, "main_input"),
    joined = inp_spec(concat_block, c(input_a = "tower_a", input_b = "tower_b")),
    output = inp_spec(output_block, "joined")
  ),
  mode = "classification"
)
```

::: {.notes}
Objetivo: mostrar que un grafo sigue encajando en tidymodels.

Qué decir:
- Con esto tenemos un grafo no lineal sin complicarnos con múltiples entradas desde la receta.
- Si tu arquitectura es una cadena, el API secuencial suele ser más simple; funcional es para “grafos”: ramas, uniones, skip connections.
- Si preguntan por multi-entrada: define varios bloques `layer_input()` y conéctalos por nombre con `inp_spec()`.

Transición: Siguiente: construir el workflow y ajustar con `tune_grid()`.
:::

---

## Workflow de Rotación (alto nivel)

```r
library(tidymodels)
library(modeldata)

data(attrition)

attr_split <- initial_split(attrition, strata = Attrition)
attr_train <- training(attr_split)

attr_rec <- recipe(Attrition ~ ., data = attr_train) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

attr_spec <- attrition_towers(
  tower_a_units = tune(),
  tower_b_units = tune(),
  fit_epochs    = 20
) |>
  set_engine("keras")

attr_wf <- workflow() |>
  add_recipe(attr_rec) |>
  add_model(attr_spec)
```

::: {.notes}
Decir:
- Mismo patron: receta + spec + workflow + tune_grid().
- Asegurate de que el outcome sea factor y los niveles sean estables.

Transicion: Ahora: resultados en CV y seleccion.
:::
---

## Resultados (Rotación): CV — Tabla

::: {.notes}
Decir:
- Son estimaciones de validacion cruzada (media + error estandar). Se selecciona con CV, no con el test.
- Busca una zona estable: buena AUC y poca sensibilidad al tamano de las torres.

Transicion: Ahora el mapa de calor para ver sensibilidad.
:::
```{r}
#| label: attrition-results-cv-table-es
#| echo: false
#| results: asis
#| message: false
#| warning: false
attrition_rds <- file.path("results", "attrition_res.rds")

if (file.exists(attrition_rds)) {
    attr_res <- readRDS(attrition_rds)
    suppressPackageStartupMessages({
        library(dplyr)
        library(tune)
    })

    best_tbl <- show_best(attr_res, metric = "roc_auc", n = 5)
    best_tbl <- best_tbl |>
        dplyr::select(tower_a_units, tower_b_units, mean, std_err) |>
        dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 4)))
    cat(knitr::kable(best_tbl, format = "html"), "\n\n")
} else {
    cat("Todavía no hay resultados en caché. Renderiza una vez para calcularlos.\n")
}
```

---

## Resultados (Rotación): CV — Gráfico

```{r}
#| label: attrition-results-cv-plot-es
#| echo: false
#| message: false
#| warning: false
attrition_rds <- file.path("results", "attrition_res.rds")

if (file.exists(attrition_rds)) {
    attr_res <- readRDS(attrition_rds)
    suppressPackageStartupMessages({
        library(dplyr)
        library(ggplot2)
        library(tune)
    })

    p_heat <- collect_metrics(attr_res) |>
        filter(.metric == "roc_auc") |>
        ggplot(aes(x = factor(tower_a_units), y = factor(tower_b_units), fill = mean)) +
        geom_tile() +
        scale_fill_gradient(low = "#f0f1f2", high = "#BF281B") +
        labs(x = "tower_a_units", y = "tower_b_units", fill = "media roc_auc", title = "Media roc_auc (CV)")
    print(p_heat)
} else {
    cat("Todavía no hay resultados en caché. Renderiza una vez para calcularlos.\n")
}
```

::: {.notes}
Objetivo: ver si una torre importa más que la otra.

Qué decir:
- Un mapa “plano” sugiere poca sensibilidad: elige simple.
- Un pico claro sugiere que merece la pena acotar la búsqueda alrededor de esa zona.
Transición: Siguiente: reportamos una sola vez en test.
:::

---

## Resultados (Rotación): Test — Tabla

```{r}
#| label: attrition-results-test-table-es
#| echo: false
#| results: asis
#| message: false
#| warning: false
attrition_final_rds <- file.path("results", "attrition_final.rds")

if (file.exists(attrition_final_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
    })

    final <- readRDS(attrition_final_rds)

    metrics_tbl <- final$metrics |>
        dplyr::select(.metric, .estimate) |>
        dplyr::mutate(.estimate = round(.estimate, 4))
    cat(knitr::kable(metrics_tbl, format = "html"), "\n\n")
} else {
    cat("Todavía no hay resultados finales en caché.\n")
}
```

::: {.notes}
Objetivo: reportar el rendimiento final una sola vez.

Qué decir:
- La selección se hizo con CV; el test es confirmación final.
- En conjuntos desbalanceados, ROC AUC suele ser más informativa que accuracy.
Transición: Siguiente: una curva ROC como resumen visual.
:::

---

## Resultados (Rotación): Test — Gráfico

```{r}
#| label: attrition-results-test-plot-es
#| echo: false
#| message: false
#| warning: false
attrition_final_rds <- file.path("results", "attrition_final.rds")

if (file.exists(attrition_final_rds)) {
    suppressPackageStartupMessages({
        library(ggplot2)
    })

    final <- readRDS(attrition_final_rds)

    if (!is.null(final$roc)) {
        p_roc <- ggplot(final$roc, aes(x = 1 - specificity, y = sensitivity)) +
            geom_line(linewidth = 1) +
            geom_abline(linetype = "dashed", alpha = 0.5) +
            coord_equal() +
            labs(x = "Tasa falsos positivos", y = "Tasa verdaderos positivos", title = "Curva ROC (test)")
        print(p_roc)
    } else {
        cat("No hay curva ROC disponible.\n")
    }
} else {
    cat("Todavía no hay resultados finales en caché.\n")
}
```

::: {.notes}
Decir:
- Cuanto mas lejos de la diagonal, mejor la discriminacion.
- Si los resultados son parecidos, elige la arquitectura mas simple dentro de la incertidumbre.

Transicion: Cerramos con consejos practicos.
:::
---

## Consejos Prácticos

- Empieza con `compile_keras_grid()` antes de ajustar
- Mantén bloques pequeños y testeables
- Pocos hiperparámetros importantes > muchos microajustes

::: {.notes}
Decir:
- Si algo falla en el grafo funcional, reduce a 2-3 bloques para aislar el error.
- Bloques pequenos y testeables; pocos hiperparametros importantes.

Transicion: Cierro con ventajas/limitaciones y hoja de ruta.
:::
---

## Ventajas y Limitaciones

### Ventajas
1.  **Paridad**: Usa `recipes`, `workflows` y `tune` igual que con `glmnet`.
2.  **Reproducibilidad**: La topología es código. Control de versiones total.
3.  **Extensibilidad**: Envuelve *cualquier* capa Keras (Transformers, RNNs) en un bloque.

### Limitaciones
-   **Bloques manuales**: no hay bloques tipo “ResNet50” prehechos (hay que definirlos).
-   **Funciones de pérdida**: una integración profunda de pérdidas personalizadas requiere algo de código repetitivo.

::: {.notes}
Decir:
- El trabajo esta en definir buenos bloques; despues, el flujo es familiar.
- No hay bloques preconstruidos (ej. ResNet50), pero cualquier capa Keras se puede envolver.

Transicion: Terminamos con hoja de ruta y abrimos preguntas.
:::
---

## Hoja de Ruta

- 📦 **Plantillas**: Plantillas para arquitecturas de alto nivel de uso común.
- 🔧 **Ajuste fino (fine-tuning)**: Flujos para adaptar modelos preentrenados.

### Recursos
- 📦 **Paquete**: [github.com/davidrsch/kerasnip](https://github.com/davidrsch/kerasnip)
- 📑 **Docs**: [davidrsch.github.io/kerasnip](https://davidrsch.github.io/kerasnip/)

::: {.notes}
Objetivo: cerrar limpio y abrir turno de preguntas.

Qué decir:
- Resumen: bloques → spec → workflow → ajuste.
- Invita a feedback y contribuciones.
- Deja tiempo para preguntas.

Transición: Deja de compartir y pasa a Q&A.
:::

:::
