---
format:
  revealjs:
    slide-level: 2
    slide-number: true
    preview-links: auto
    footer: "kerasnip v0.1.0.900"
    mermaid:
      theme: neutral
    width: 1280
    height: 720
    margin: 0.05
    min-scale: 0.2
    max-scale: 2.0
execute:
  freeze: auto
  cache: true
---

```{r}
#| label: madrid-2-2026-cache-results
#| include: false
#| message: false
#| warning: false

# Compute results once, cache them on disk, and reuse across EN/ES renders.
#
# Quarto executes this document with working directory = `presentations/madrid-2-2026/`,
# so cache files should live in `presentations/madrid-2-2026/results/` (i.e. `results/` here).
results_dir <- "results"
old_results_dirs <- c(
    file.path("..", "..", "madrid-2026", "results"),
    file.path("madrid-2026", "results")
)
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

# Migration:
# - Before the folder move, cached results lived at `madrid-2026/results/`.
# - Older versions could also write nested `madrid-2026/results/` inside the project dir.
for (old_results_dir in old_results_dirs) {
    if (!dir.exists(old_results_dir)) {
        next
    }

    old_rds <- list.files(
        old_results_dir,
        pattern = "\\.rds$",
        full.names = TRUE
    )
    if (length(old_rds) > 0) {
        file.copy(old_rds, results_dir, overwrite = FALSE)
    }
}

diamonds_rds <- file.path(results_dir, "diamonds_res.rds")
diamonds_final_rds <- file.path(results_dir, "diamonds_final.rds")
attrition_rds <- file.path(results_dir, "attrition_res.rds")
attrition_final_rds <- file.path(results_dir, "attrition_final.rds")

set.seed(20260203)

can_run <- function(...) {
    all(vapply(list(...), requireNamespace, logical(1), quietly = TRUE))
}

if (can_run("tidymodels", "kerasnip", "keras3")) {
    suppressPackageStartupMessages({
        library(tidymodels)
        library(kerasnip)
        library(keras3)
    })

    data(diamonds, package = "ggplot2")

    input_block <- function(model, input_shape) {
        keras_model_sequential(input_shape = input_shape)
    }

    dense_block <- function(model, units = 128, dropout = 0.0) {
        model |>
            layer_dense(units = units, activation = "relu") |>
            layer_dropout(rate = dropout)
    }

    output_block <- function(model) {
        model |> layer_dense(units = 1)
    }

    create_keras_sequential_spec(
        model_name = "diamond_mlp",
        layer_blocks = list(
            input = input_block,
            body = dense_block,
            output = output_block
        ),
        mode = "regression"
    )

    diamonds_split <- initial_split(diamonds, strata = price)
    diamonds_train <- training(diamonds_split)

    diamonds_rec <- recipe(price ~ ., data = diamonds_train) |>
        step_dummy(all_nominal_predictors()) |>
        step_zv(all_predictors()) |>
        step_normalize(all_numeric_predictors())

    spec <- diamond_mlp(
        num_body = tune(),
        body_units = tune(),
        body_dropout = tune(),
        fit_epochs = 5
    ) |>
        set_engine("keras")

    wf <- workflow() |>
        add_recipe(diamonds_rec) |>
        add_model(spec)

    params <- extract_parameter_set_dials(wf) |>
        update(
            num_body = dials::num_terms(c(1, 3)),
            body_units = dials::hidden_units(c(32, 128)),
            body_dropout = dials::dropout(c(0, 0.3))
        )

    grid <- grid_latin_hypercube(params, size = 8)
    folds <- vfold_cv(diamonds_train, v = 3, strata = price)

    if (!file.exists(diamonds_rds)) {
        res <- tune_grid(
            wf,
            resamples = folds,
            grid = grid,
            control = control_grid()
        )
        saveRDS(res, diamonds_rds)
    } else {
        res <- readRDS(diamonds_rds)
    }

    if (!file.exists(diamonds_final_rds)) {
        best <- select_best(res, metric = "rmse")
        final_wf <- finalize_workflow(wf, best)
        lf <- last_fit(
            final_wf,
            split = diamonds_split,
            metrics = metric_set(rmse, mae, rsq)
        )

        diamonds_final <- list(
            best = best,
            metrics = collect_metrics(lf),
            preds = collect_predictions(lf)
        )
        saveRDS(diamonds_final, diamonds_final_rds)
    }
}

if (can_run("tidymodels", "modeldata", "kerasnip", "keras3")) {
    suppressPackageStartupMessages({
        library(tidymodels)
        library(modeldata)
        library(kerasnip)
        library(keras3)
    })

    data(attrition)

    # Correct functional signatures per kerasnip docs:
    input_block <- function(input_shape) {
        layer_input(shape = input_shape, name = "features")
    }

    dense_block <- function(tensor, units = 64, dropout = 0.0) {
        tensor |>
            layer_dense(units = units, activation = "relu") |>
            layer_dropout(rate = dropout)
    }

    concat_block <- function(input_a, input_b) {
        layer_concatenate(list(input_a, input_b))
    }

    output_block <- function(tensor, num_classes) {
        tensor |> layer_dense(units = num_classes, activation = "softmax")
    }

    create_keras_functional_spec(
        model_name = "attrition_towers",
        layer_blocks = list(
            main_input = input_block,
            tower_a = inp_spec(dense_block, "main_input"),
            tower_b = inp_spec(dense_block, "main_input"),
            joined = inp_spec(
                concat_block,
                c(input_a = "tower_a", input_b = "tower_b")
            ),
            output = inp_spec(output_block, "joined")
        ),
        mode = "classification"
    )

    split <- initial_split(attrition, strata = Attrition)
    train <- training(split)

    rec <- recipe(Attrition ~ ., data = train) |>
        step_zv(all_predictors()) |>
        step_dummy(all_nominal_predictors()) |>
        step_normalize(all_numeric_predictors())

    spec <- attrition_towers(
        tower_a_units = tune(),
        tower_b_units = tune(),
        fit_epochs = 5
    ) |>
        set_engine("keras")

    wf <- workflow() |>
        add_recipe(rec) |>
        add_model(spec)

    params <- extract_parameter_set_dials(wf) |>
        update(
            tower_a_units = dials::hidden_units(c(16, 128)),
            tower_b_units = dials::hidden_units(c(16, 128))
        )

    grid <- grid_latin_hypercube(params, size = 8)
    folds <- vfold_cv(train, v = 3, strata = Attrition)

    if (!file.exists(attrition_rds)) {
        res <- tune_grid(
            wf,
            resamples = folds,
            grid = grid,
            metrics = metric_set(roc_auc, accuracy)
        )
        saveRDS(res, attrition_rds)
    } else {
        res <- readRDS(attrition_rds)
    }

    if (!file.exists(attrition_final_rds)) {
        best <- select_best(res, metric = "roc_auc")
        final_wf <- finalize_workflow(wf, best)
        lf <- last_fit(
            final_wf,
            split = split,
            metrics = metric_set(roc_auc, accuracy)
        )

        preds <- collect_predictions(lf)
        event_level <- levels(preds$Attrition)[1]
        preferred <- paste0(".pred_", event_level)
        pos_col <- if (preferred %in% names(preds)) {
            preferred
        } else {
            pcols <- grep("^\\.pred_", names(preds), value = TRUE)
            pcols <- setdiff(pcols, ".pred_class")
            pcols[1]
        }

        attrition_final <- list(
            best = best,
            metrics = collect_metrics(lf),
            preds = preds,
            roc = if (!is.na(pos_col)) {
                roc_curve(preds, truth = Attrition, !!rlang::sym(pos_col))
            } else {
                NULL
            }
        )
        saveRDS(attrition_final, attrition_final_rds)
    }
}
```

::: {.content-visible when-profile="english"}

## kerasnip: A Bridge Between 'keras' and 'tidymodels' {.title-slide background-color="#f0f1f2"}

### Madrid R Group | February 2026
**kerasnip v0.1.0.900**

Docs: [davidrsch.github.io/kerasnip](https://davidrsch.github.io/kerasnip/) · Repo: [github.com/davidrsch/kerasnip](https://github.com/davidrsch/kerasnip)

::: {.notes}
Open with energy. Ask: "Who loves tidymodels?" then "Who has fought with Keras architectures?"

Say:
- Keras and tidymodels are like oil and water, kerasnip is the emulsifier.
- Today: solving the architectural tuning problem.
:::

---

## Today

- Why tuning Keras inside tidymodels is hard
- What `kerasnip` generates and why it matters
- Sequential workflow: structural tuning with `num_{block}`
- Functional workflow: branch/merge graphs with `inp_spec()`

::: {.notes}
Keep this slide to 20 seconds. It is orientation, not content.

Say:
- We will go from problem to working code quickly.
- First half: sequential API. Second half: functional graphs.
- There is a debug helper at the end that is worth knowing about.
:::

---

## The Landscape & The Gap

::: {.columns}
::: {.column width="50%"}
### Keras (via `keras3`)
- **Deep Learning**
- Tensor-centric & imperative
- Backend agnostic (TF, JAX, Torch)
- Arbitrary graphs of layers
:::

::: {.column width="50%"}
### tidymodels
- **Machine Learning**
- Dataframe-centric & declarative
- Unified interface (`fit`, `predict`)
- Standardized tuning
:::
:::

**The gap**: `tune_grid()` handles learning rate easily, but tuning *model structure*: depth, width, branches; is awkward. `parsnip` expects a fixed set of named arguments; Keras models are arbitrary graphs.

::: {.notes}
Analogy: Keras = architect's blank sheet; tidymodels = government form with fixed fields. kerasnip makes the forms fit the drawings.

Say:
- You can already tune learning rate with keras3 + tidymodels. The friction is topology, which live in Keras code, not in named spec arguments.
- kerasnip does not replace keras3. It generates new parsnip spec functions that use the existing keras engine.
- One sentence on why not other packages: reticulate is too low-level, luz/brulee are Torch-only, parsnip::mlp() is too fixed for architecture search.

Transition: Let us see how kerasnip solves this, one diagram.
:::

---

## Mental Model {.mermaid-wide}

<br/>

```{mermaid}
%%| fig-align: center

flowchart LR
  classDef step fill:#f5f7fa,stroke:#5c6b7a,color:#1b3a57,stroke-width:2px;
  classDef tune fill:#BF281B,stroke:#BF281B,color:#ffffff,stroke-width:2px;

  A["Blocks<br/>(plain R functions)"] --> B["Spec function<br/>create_keras_*_spec()"]
  B --> C["Workflow<br/>recipe() + workflow()"]
  C --> D["Tuning<br/>tune_grid()"]
  A -. "args → tunables<br/>num_{block} repeats" .-> B

  class A,B,C step;
  class D tune;
```

<br/>

**One rule**: if you want to tune it, make it a function argument.

::: {.notes}
Spend about 60 seconds here. This diagram is the whole talk in one image.

Say:
- You write plain R functions called blocks. Each block returns Keras layers.
- create_keras_*_spec() reads those functions and generates a new parsnip model.
- From that point on it is standard tidymodels: recipe, workflow, tune_grid.
- The dotted arrow is the key insight: block arguments automatically become tunable spec arguments, prefixed by block name. dense_block(units=128) becomes body_units. And num_{block} controls how many times a block repeats, that is depth as a hyperparameter.

If someone gets lost: blocks → spec → workflow → tuning. Everything else is detail.

Transition: Let us build one. Sequential API first, predicting diamond prices.
:::

---

## Sequential Blocks

```r
input_block <- function(model, input_shape) {
  keras_model_sequential(input_shape = input_shape)
}

dense_block <- function(model, units = 128, dropout = 0.0) {
  model |>
    layer_dense(units = units, activation = "relu") |>
    layer_dropout(rate = dropout)
}

output_block <- function(model) {
  model |> layer_dense(units = 1)
}
```

::: {.notes}
Say:
- These are plain R functions. Nothing from kerasnip here yet.
- The function signature is everything: units and dropout become tunable because they are arguments with defaults.
- dense_block is the repeatable one, num_body will control how many times it stacks.
- input and output blocks have no tunable args by design: input shape is fixed by the data, output is fixed by the task.

Point at the code while saying: "body_units comes from units here, body_dropout from dropout. The prefix body comes from the block name we will assign in the list on the next slide."

Transition: Now pass these to create_keras_sequential_spec and we get a parsnip model.
:::

---

## Generate Spec + Structural Tuning

::: {.columns}
::: {.column width="50%"}
**Register the spec**

```r
create_keras_sequential_spec(
  model_name = "diamond_mlp",
  layer_blocks = list(
    input  = input_block,
    body   = dense_block,  # repeatable
    output = output_block
  ),
  mode = "regression"
)
```

This creates `diamond_mlp()`, a standard parsnip model function.
:::

::: {.column width="50%"}
**Treat topology as a hyperparameter**

```r
spec <- diamond_mlp(
  num_body     = tune(),  # depth: 1, 2, 3…
  body_units   = tune(),  # width: 32, 64, 128…
  body_dropout = tune(),
  fit_epochs   = 20
) |>
  set_engine("keras")
```

`num_body` and `body_units` are auto-generated from the block name and its arguments.
:::
:::

::: {.notes}
Say:
- Left: one call, one new parsnip model. That is all kerasnip does structurally.
- Right: now depth is tune() just like any other hyperparameter. This is what was not possible before.
- Point at num_body: this controls how many times dense_block stacks. Put tune() here and you are searching architecture space.
- Point at body_units: the body_ prefix comes from naming the block "body" in the list on the left.

Transition: Now drop it into a standard workflow.
:::

---

## Diamonds Workflow

```r
library(tidymodels);  library(kerasnip)
data(diamonds, package = "ggplot2")

diamonds_split <- initial_split(diamonds, strata = price)
diamonds_train <- training(diamonds_split)

diamonds_rec <- recipe(price ~ ., data = diamonds_train) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

wf <- workflow() |> add_recipe(diamonds_rec) |> add_model(spec)

params <- extract_parameter_set_dials(wf) |>
  update(
    num_body     = dials::num_terms(c(1, 4)),
    body_units   = dials::hidden_units(c(32, 256)),
    body_dropout = dials::dropout(c(0, 0.4))
  )

grid  <- grid_latin_hypercube(params, size = 20)
folds <- vfold_cv(diamonds_train, v = 5, strata = price)
res   <- tune_grid(wf, resamples = folds, grid = grid)
```

::: {.notes}
Say:
- This is identical to any other tidymodels workflow. That is the point.
- Recipe must output numeric matrices: step_dummy + step_normalize are required for Keras.
- extract_parameter_set_dials discovers tunables automatically, you just update the ranges.
- Practical tip: start with size = 8 when testing a new architecture. Scale up once you know it builds correctly.

Transition: Let us look at what came out.
:::

---

## Results (Diamonds): CV — Table

```{r}
#| label: diamonds-results-cv-table
#| echo: false
#| results: asis
#| message: false
#| warning: false
diamonds_rds <- file.path("results", "diamonds_res.rds")

if (file.exists(diamonds_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
        library(tidymodels)
    })
    res <- readRDS(diamonds_rds)
    best_tbl <- show_best(res, metric = "rmse", n = 5)
    best_tbl <- best_tbl |>
        dplyr::select(num_body, body_units, body_dropout, mean, std_err) |>
        dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3)))
    cat(knitr::kable(best_tbl, format = "html"), "\n\n")
} else {
    cat("No cached results found yet. Render once to compute them.\n")
}
```

::: {.notes}
Keep this slide to 30 seconds then move to the plot.

Say:
- These are cross-validation estimates. We select on CV, never on the test set.
- Look for a stable region: good mean RMSE and small std_err. That is the architecture to carry forward.
- Pick the simplest within 1 std_err of the best, not necessarily the top row.
:::

---

## Results (Diamonds): CV — Plot

```{r}
#| label: diamonds-results-cv-plot
#| echo: false
#| message: false
#| warning: false
diamonds_rds <- file.path("results", "diamonds_res.rds")

if (file.exists(diamonds_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
        library(ggplot2)
        library(tidymodels)
    })
    res <- readRDS(diamonds_rds)
    p <- collect_metrics(res) |>
        filter(.metric == "rmse") |>
        ggplot(aes(
            x = num_body,
            y = mean,
            color = factor(body_units),
            group = interaction(body_units, body_dropout)
        )) +
        geom_point() +
        geom_line(alpha = 0.35) +
        facet_wrap(~body_dropout) +
        labs(
            x = "num_body (depth)",
            y = "CV RMSE (mean)",
            color = "body_units",
            title = "Cross-validation trends across depth / width / dropout"
        )
    print(p)
} else {
    cat("No cached results found yet. Render once to compute them.\n")
}
```

::: {.notes}
Say:
- If lines flatten as num_body increases, going deeper is not buying anything, pick the shallower model.
- Each facet is a dropout value: look for a panel where the spread between width options is small. That is a robust region.

Transition: One-time test confirmation.
:::

---

## Results (Diamonds): Test — Table

```{r}
#| label: diamonds-results-test-table
#| echo: false
#| results: asis
#| message: false
#| warning: false
diamonds_final_rds <- file.path("results", "diamonds_final.rds")

if (file.exists(diamonds_final_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
    })
    final <- readRDS(diamonds_final_rds)
    metrics_tbl <- final$metrics |>
        dplyr::select(.metric, .estimate) |>
        dplyr::mutate(.estimate = round(.estimate, 4))
    cat(knitr::kable(metrics_tbl, format = "html"), "\n\n")
} else {
    cat("No cached final results found yet.\n")
}
```

::: {.notes}
Say:
- Test is for reporting only. The architecture was chosen with CV, this is just confirmation.
- If test is much worse than CV, suspect data leakage or an unstable architecture.
:::

---

## Results (Diamonds): Test — Plot

```{r}
#| label: diamonds-results-test-plot
#| echo: false
#| message: false
#| warning: false
diamonds_final_rds <- file.path("results", "diamonds_final.rds")

if (file.exists(diamonds_final_rds)) {
    suppressPackageStartupMessages({
        library(ggplot2)
    })
    final <- readRDS(diamonds_final_rds)
    p_test <- final$preds |>
        ggplot(aes(x = price, y = .pred)) +
        geom_point(alpha = 0.25) +
        geom_abline(slope = 1, intercept = 0, linewidth = 0.8) +
        coord_fixed() +
        labs(
            x = "Truth (price)",
            y = "Predicted",
            title = "Final model on test set"
        )
    print(p_test)
} else {
    cat("No cached final results found yet.\n")
}
```

::: {.notes}
Say:
- Points near the diagonal are good predictions. Systematic deviation below at high prices means the model underestimates expensive diamonds, expected, as those are rare in training data.

Transition: Before moving to functional, one slide on a tool that saves a lot of debugging pain.
:::

---

## Debugging: `compile_keras_grid()` {.compact .table-compact}

```r
prep_rec <- prep(diamonds_rec, training = diamonds_train)
juiced <- juice(prep_rec)
x <- juiced |> select(-price)
y <- juiced$price
modified_grid <- tibble(
    num_body = c(3, 6),
    body_units = c(-10, 128),
    body_dropout = c(0.2, 0.2)
)

dbg <- compile_keras_grid(spec, modified_grid, x, y)
dbg
```

```{r}
#| label: diamonds-debug
#| echo: false
#| message: false
#| warning: false
suppressPackageStartupMessages({
    library(dplyr)
    library(tibble)
    library(purrr)
    library(tidymodels)
    library(kerasnip)
    library(keras3)
})

if (
    !exists("diamond_mlp", mode = "function") ||
        !kerasnip::model_exists("diamond_mlp")
) {
    input_block <- function(model, input_shape) {
        keras_model_sequential(input_shape = input_shape)
    }
    dense_block <- function(model, units = 128, dropout = 0.0) {
        model |>
            layer_dense(units = units, activation = "relu") |>
            layer_dropout(rate = dropout)
    }
    output_block <- function(model) model |> layer_dense(units = 1)
    create_keras_sequential_spec(
        model_name = "diamond_mlp",
        layer_blocks = list(
            input = input_block,
            body = dense_block,
            output = output_block
        ),
        mode = "regression"
    )
}

prep_rec <- prep(diamonds_rec, training = diamonds_train)
juiced <- juice(prep_rec)
x <- juiced |> select(-price)
y <- juiced$price

test_grid <- tibble(
    num_body = c(3, 6),
    body_units = c(-10, 128),
    body_dropout = c(0.2, 0.2)
)

dbg <- compile_keras_grid(
    spec = diamond_mlp(
        num_body = tune(),
        body_units = tune(),
        body_dropout = tune(),
        fit_epochs = 5
    ) |>
        set_engine("keras"),
    grid = test_grid,
    x = x,
    y = y
)

dbg
```

::: {.notes}
Say:
- Before running tune_grid on 20 configurations, run this on one to confirm the model actually builds correctly.
- The error column shows which grid rows failed and why, catch this before wasting compute.
- extract_valid_grid() filters to successful builds; inform_errors() summarizes failures across the whole grid.
- Rule: always compile before you tune.

Transition: Same workflow pattern, now with a non-linear graph. Functional API.
:::

---

## Functional API: Blocks + Graph

::: {.columns}
::: {.column width="50%"}
**Blocks receive tensors, not models**

```r
input_block <- function(input_shape) {
  layer_input(shape = input_shape,
              name = "features")
}

dense_block <- function(tensor,
                        units = 64,
                        dropout = 0.0) {
  tensor |>
    layer_dense(units = units,
                activation = "relu") |>
    layer_dropout(rate = dropout)
}

concat_block <- function(input_a, input_b) {
  layer_concatenate(list(input_a, input_b))
}

output_block <- function(tensor, num_classes) {
  tensor |>
    layer_dense(units = num_classes,
                activation = "softmax")
}
```
:::

::: {.column width="50%"}
**`inp_spec()` wires outputs to inputs by name**

```r
create_keras_functional_spec(
  model_name = "attrition_towers",
  layer_blocks = list(
    main_input = input_block,
    tower_a    = inp_spec(
                   dense_block,
                   "main_input"),
    tower_b    = inp_spec(
                   dense_block,
                   "main_input"),
    joined     = inp_spec(
                   concat_block,
                   c(input_a = "tower_a",
                     input_b = "tower_b")),
    output     = inp_spec(
                   output_block,
                   "joined")
  ),
  mode = "classification"
)
```
:::
:::

::: {.notes}
Say:
- Sequential: blocks receive a model and stack layers onto it. Functional: blocks receive tensors and return tensors. That is the only structural difference.
- inp_spec(func, inputs) is just labeling which upstream output feeds which argument, like patch cables on a modular synth.
- This graph: one input splits into two parallel towers, they concatenate, then output. Non-linear, but the workflow stays completely identical.
- Use sequential for a straight chain. Use functional when you need branches, merges, or skip connections.
- tower_a_units and tower_b_units are auto-generated by the same prefix convention as before.

Transition: Now the workflow, you will recognize every line.
:::

---

## Attrition Workflow

```r
library(tidymodels);  library(modeldata)
data(attrition)

attr_split <- initial_split(attrition, strata = Attrition)
attr_train <- training(attr_split)

attr_rec <- recipe(Attrition ~ ., data = attr_train) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

attr_spec <- attrition_towers(
  tower_a_units = tune(),
  tower_b_units = tune(),
  fit_epochs    = 20
) |> set_engine("keras")

attr_wf <- workflow() |> add_recipe(attr_rec) |> add_model(attr_spec)

params <- extract_parameter_set_dials(attr_wf) |>
  update(tower_a_units = dials::hidden_units(c(16, 128)),
         tower_b_units = dials::hidden_units(c(16, 128)))

res <- tune_grid(attr_wf,
                 resamples = vfold_cv(attr_train, v = 3, strata = Attrition),
                 grid      = grid_latin_hypercube(params, size = 12))
```

::: {.notes}
Say:
- Identical pattern to the sequential example. That is the whole point of kerasnip, the graph complexity is completely hidden.
- Make sure the outcome is a factor with stable levels.

Transition: Results.
:::

---

## Results (Attrition): CV — Table

```{r}
#| label: attrition-results-cv-table
#| echo: false
#| results: asis
#| message: false
#| warning: false
attrition_rds <- file.path("results", "attrition_res.rds")

if (file.exists(attrition_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
        library(tune)
    })
    attr_res <- readRDS(attrition_rds)
    best_tbl <- show_best(attr_res, metric = "roc_auc", n = 5)
    best_tbl <- best_tbl |>
        dplyr::select(tower_a_units, tower_b_units, mean, std_err) |>
        dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 4)))
    cat(knitr::kable(best_tbl, format = "html"), "\n\n")
} else {
    cat("No cached results found yet. Render once to compute them.\n")
}
```

::: {.notes}
Keep this slide to 30 seconds then move to the heatmap.

Say:
- Same reading as before: mean ROC AUC plus std_err. Select on CV.
- Pick the simplest architecture within 1 std_err of the best.
:::

---

## Results (Attrition): CV — Plot

```{r}
#| label: attrition-results-cv-plot
#| echo: false
#| message: false
#| warning: false
attrition_rds <- file.path("results", "attrition_res.rds")

if (file.exists(attrition_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
        library(ggplot2)
        library(tune)
    })
    attr_res <- readRDS(attrition_rds)
    p_heat <- collect_metrics(attr_res) |>
        filter(.metric == "roc_auc") |>
        ggplot(aes(
            x = factor(tower_a_units),
            y = factor(tower_b_units),
            fill = mean
        )) +
        geom_tile() +
        scale_fill_gradient(low = "#f0f1f2", high = "#BF281B") +
        labs(
            x = "tower_a_units",
            y = "tower_b_units",
            fill = "mean roc_auc",
            title = "CV mean roc_auc"
        )
    print(p_heat)
} else {
    cat("No cached results found yet. Render once to compute them.\n")
}
```

::: {.notes}
Say:
- A flat heatmap means the architecture is robust, tower size does not matter much, pick the smallest.
- A sharp peak means the model is sensitive to tower sizes, worth narrowing the search around that region in a second pass.

Transition: Final test confirmation.
:::

---

## Results (Attrition): Test — Table

```{r}
#| label: attrition-results-test-table
#| echo: false
#| results: asis
#| message: false
#| warning: false
attrition_final_rds <- file.path("results", "attrition_final.rds")

if (file.exists(attrition_final_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
    })
    final <- readRDS(attrition_final_rds)
    metrics_tbl <- final$metrics |>
        dplyr::select(.metric, .estimate) |>
        dplyr::mutate(.estimate = round(.estimate, 4))
    cat(knitr::kable(metrics_tbl, format = "html"), "\n\n")
} else {
    cat("No cached final results found yet.\n")
}
```

::: {.notes}
Say:
- On imbalanced datasets like attrition, ROC AUC is more informative than accuracy, accuracy can look fine just by predicting the majority class.
- Again: test is for reporting, not for deciding.
:::

---

## Results (Attrition): Test — Plot

```{r}
#| label: attrition-results-test-plot
#| echo: false
#| message: false
#| warning: false
attrition_final_rds <- file.path("results", "attrition_final.rds")

if (file.exists(attrition_final_rds)) {
    suppressPackageStartupMessages({
        library(ggplot2)
    })
    final <- readRDS(attrition_final_rds)
    if (!is.null(final$roc)) {
        p_roc <- ggplot(final$roc, aes(x = 1 - specificity, y = sensitivity)) +
            geom_line(linewidth = 1) +
            geom_abline(linetype = "dashed", alpha = 0.5) +
            coord_equal() +
            labs(
                x = "False positive rate",
                y = "True positive rate",
                title = "ROC curve (test set)"
            )
        print(p_roc)
    } else {
        cat("ROC curve not available.\n")
    }
} else {
    cat("No cached final results found yet.\n")
}
```

::: {.notes}
Say:
- Further from the diagonal = better discrimination.
- When two architectures give similar curves, always pick the simpler one.

Transition: Close with the practical summary.
:::

---

## Practical Tips & Trade-offs

::: {.columns}
::: {.column width="50%"}
**To avoid pain**

- Always run `compile_keras_grid()` before tuning
- Keep blocks small and testable, isolate errors with 2–3 blocks
- Tune structure (depth/width) first; training hyperparameters later
- Keep search spaces reasonable, depth × width explodes fast

:::

::: {.column width="50%"}
**Why it is worth it**

1. **Workflow parity** — `recipes`, `workflows`, `tune` work exactly like with `glmnet`
2. **Reproducibility** — topology is code; version-control your architecture
3. **Extensibility** — wrap any Keras layer (Transformers, RNNs) into a block

<br/>

:::
:::

::: {.notes}
Goal: leave people with a clear answer to "should I try this?"

Say:
- The learning curve is writing good blocks. After that, everything is standard tidymodels.
- If you already use tidymodels, the only new concept is defining blocks and calling create_keras_*_spec().
- Package is on GitHub, contributions and feedback welcome.

Stop sharing screen and open Q&A.
:::

---

## Resources

- 📦 [github.com/davidrsch/kerasnip](https://github.com/davidrsch/kerasnip)
- 📑 [davidrsch.github.io/kerasnip](https://davidrsch.github.io/kerasnip/)
- 📑 [https://davidrsch.github.io/kerasnip_presentations/presentations/madrid-2-2026/)


:::

::: {.content-visible when-profile="spanish"}

## kerasnip: Un Puente entre 'keras' y 'tidymodels' {.title-slide background-color="#f0f1f2"}

### Grupo de Usuarios de R de Madrid | Febrero 2026
**kerasnip v0.1.0.900**

Docs: [davidrsch.github.io/kerasnip](https://davidrsch.github.io/kerasnip/) · Repo: [github.com/davidrsch/kerasnip](https://github.com/davidrsch/kerasnip)

::: {.notes}
Empezar con energia. Preguntar: "¿Quién usa tidymodels?" y "¿Quién ha peleado con arquitecturas Keras?"

Decir:
- Keras y tidymodels son como aceite y agua; kerasnip es el emulsionante.
- Hoy: resolver el problema del ajuste de arquitectura.
:::

---

## Hoy

- Por qué es difícil ajustar modelos Keras dentro de tidymodels
- Qué genera `kerasnip` y por qué importa
- Flujo secuencial: ajuste estructural con `num_{block}`
- Flujo funcional: grafos con ramas y uniones con `inp_spec()`

::: {.notes}
Mantener esta diapositiva en 20 segundos. Es orientación, no contenido.

Decir:
- Pasamos del problema al código funcional rápidamente.
- Primera parte: API secuencial. Segunda parte: grafos funcionales.
- Al final hay una herramienta de depuración que vale la pena conocer.
:::

---

## El Panorama y la Brecha

::: {.columns}
::: {.column width="50%"}
### Keras (vía `keras3`)
- **Aprendizaje profundo**
- Pensado para tensores, imperativo
- Agnóstico al backend (TF, JAX, Torch)
- Grafos arbitrarios de capas
:::

::: {.column width="50%"}
### tidymodels
- **Aprendizaje automático**
- Pensado para data frames, declarativo
- Interfaz unificada (`fit`, `predict`)
- Ajuste de hiperparámetros estandarizado
:::
:::

**La brecha**: `tune_grid()` maneja el learning rate sin problema, pero ajustar la *estructura* es imposible. `parsnip` espera argumentos fijos; los modelos Keras son grafos arbitrarios.

::: {.notes}
Analogia: Keras = mesa de arquitecto (papel en blanco); tidymodels = formulario con casillas fijas. kerasnip adapta los formularios a los dibujos.

Decir:
- Ya se puede ajustar el learning rate con keras3 + tidymodels. La fricción está en la topología, profundidad y ancho viven en código Keras, no en argumentos con nombre del spec.
- kerasnip no reemplaza keras3. Genera nuevas funciones spec de parsnip que usan el engine keras existente.
- Una frase sobre por qué no otros paquetes: reticulate es demasiado bajo nivel, luz/brulee son solo Torch, parsnip::mlp() es demasiado fijo para búsqueda arquitectónica.

Transición: Veamos cómo lo resuelve kerasnip, un solo diagrama.
:::

---

## Modelo Mental {.mermaid-wide}

<br/>

```{mermaid}
%%| fig-align: center

flowchart LR
  classDef step fill:#f5f7fa,stroke:#5c6b7a,color:#1b3a57,stroke-width:2px;
  classDef tune fill:#BF281B,stroke:#BF281B,color:#ffffff,stroke-width:2px;

  A["Bloques<br/>(funciones R simples)"] --> B["Función spec<br/>create_keras_*_spec()"]
  B --> C["Workflow<br/>recipe() + workflow()"]
  C --> D["Ajuste<br/>tune_grid()"]
  A -. "args → ajustables<br/>num_{block} repite" .-> B

  class A,B,C step;
  class D tune;
```

<br/>

**Una regla**: si quieres ajustarlo, hazlo argumento de la función.

::: {.notes}
Dedicar unos 60 segundos aquí. Este diagrama es toda la charla en una imagen.

Decir:
- Escribes funciones R simples llamadas bloques. Cada bloque devuelve capas Keras.
- create_keras_*_spec() lee esas funciones y genera un nuevo modelo parsnip.
- A partir de ahí es tidymodels estándar: recipe, workflow, tune_grid.
- La flecha punteada es la idea clave: los argumentos de los bloques se convierten automáticamente en argumentos ajustables del spec, prefijados con el nombre del bloque. dense_block(units=128) se convierte en body_units. Y num_{block} controla cuántas veces se repite un bloque, eso es la profundidad como hiperparámetro.

Si alguien se pierde: bloques → spec → workflow → ajuste. Todo lo demás es detalle.

Transición: Construyamos uno. Primero el API secuencial, predecir precios de diamantes.
:::

---

## Bloques Secuenciales

```r
input_block <- function(model, input_shape) {
  keras_model_sequential(input_shape = input_shape)
}

dense_block <- function(model, units = 128, dropout = 0.0) {
  model |>
    layer_dense(units = units, activation = "relu") |>
    layer_dropout(rate = dropout)
}

output_block <- function(model) {
  model |> layer_dense(units = 1)
}
```

::: {.notes}
Decir:
- Estas son funciones R simples. Nada de kerasnip aquí todavía.
- La firma de la función es todo: units y dropout se vuelven ajustables porque son argumentos con valores por defecto.
- dense_block es el repetible, num_body controlará cuántas veces se apila.
- input y output no tienen argumentos ajustables a propósito: la forma de entrada la fija el dato, la salida la fija la tarea.

Señalar el código mientras dices: "body_units viene de units aquí, body_dropout de dropout. El prefijo body viene del nombre que le daremos al bloque en la lista de la siguiente diapositiva."

Transición: Ahora pasamos estos bloques a create_keras_sequential_spec y obtenemos un modelo parsnip.
:::

---

## Generar Spec + Ajuste Estructural

::: {.columns}
::: {.column width="50%"}
**Registrar el spec**

```r
create_keras_sequential_spec(
  model_name = "diamond_mlp",
  layer_blocks = list(
    input  = input_block,
    body   = dense_block,  # repetible
    output = output_block
  ),
  mode = "regression"
)
```

Esto crea `diamond_mlp()`, una función de modelo parsnip estándar.
:::

::: {.column width="50%"}
**Tratar la topología como hiperparámetro**

```r
spec <- diamond_mlp(
  num_body     = tune(),  # profundidad: 1, 2, 3…
  body_units   = tune(),  # ancho: 32, 64, 128…
  body_dropout = tune(),
  fit_epochs   = 20
) |>
  set_engine("keras")
```

`num_body` y `body_units` se generan automáticamente del nombre del bloque y sus argumentos.
:::
:::

::: {.notes}
Decir:
- Izquierda: una llamada, un nuevo modelo parsnip. Eso es todo lo que hace kerasnip estructuralmente.
- Derecha: ahora la profundidad es tune() igual que cualquier otro hiperparámetro. Esto es lo que no era posible antes.
- Señalar num_body: esto controla cuántas veces se apila dense_block. Con tune() estás buscando en el espacio de arquitecturas.
- Señalar body_units: el prefijo body_ viene de llamar "body" al bloque en la lista de la izquierda.

Transición: Ahora lo metemos en un workflow estándar.
:::

---

## Workflow con Diamantes

```r
library(tidymodels);  library(kerasnip)
data(diamonds, package = "ggplot2")

diamonds_split <- initial_split(diamonds, strata = price)
diamonds_train <- training(diamonds_split)

diamonds_rec <- recipe(price ~ ., data = diamonds_train) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

wf <- workflow() |> add_recipe(diamonds_rec) |> add_model(spec)

params <- extract_parameter_set_dials(wf) |>
  update(
    num_body     = dials::num_terms(c(1, 4)),
    body_units   = dials::hidden_units(c(32, 256)),
    body_dropout = dials::dropout(c(0, 0.4))
  )

grid  <- grid_latin_hypercube(params, size = 20)
folds <- vfold_cv(diamonds_train, v = 5, strata = price)
res   <- tune_grid(wf, resamples = folds, grid = grid)
```

::: {.notes}
Decir:
- Esto es idéntico a cualquier otro workflow de tidymodels. Ese es el punto.
- La receta debe producir matrices numéricas: step_dummy + step_normalize son obligatorios para Keras.
- extract_parameter_set_dials detecta los ajustables automáticamente, solo actualizas los rangos.
- Consejo práctico: empieza con size = 8 al probar una arquitectura nueva. Escala una vez que sabes que construye correctamente.

Transición: Veamos qué salió.
:::

---

## Resultados (Diamantes): CV — Tabla

```{r}
#| label: diamonds-results-cv-table-es
#| echo: false
#| results: asis
#| message: false
#| warning: false
diamonds_rds <- file.path("results", "diamonds_res.rds")

if (file.exists(diamonds_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
        library(tidymodels)
    })
    res <- readRDS(diamonds_rds)
    best_tbl <- show_best(res, metric = "rmse", n = 5)
    best_tbl <- best_tbl |>
        dplyr::select(num_body, body_units, body_dropout, mean, std_err) |>
        dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3)))
    cat(knitr::kable(best_tbl, format = "html"), "\n\n")
} else {
    cat(
        "Todavía no hay resultados en caché. Renderiza una vez para calcularlos.\n"
    )
}
```

::: {.notes}
Mantener esta diapositiva en 30 segundos y pasar al gráfico.

Decir:
- Estas son estimaciones de validación cruzada. Seleccionamos con CV, nunca con el test.
- Busca una zona estable: buen RMSE medio y std_err pequeño. Esa es la arquitectura a llevar adelante.
- Elige la más simple dentro de 1 std_err del mejor, no necesariamente la primera fila.
:::

---

## Resultados (Diamantes): CV — Gráfico

```{r}
#| label: diamonds-results-cv-plot-es
#| echo: false
#| message: false
#| warning: false
diamonds_rds <- file.path("results", "diamonds_res.rds")

if (file.exists(diamonds_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
        library(ggplot2)
        library(tidymodels)
    })
    res <- readRDS(diamonds_rds)
    p <- collect_metrics(res) |>
        filter(.metric == "rmse") |>
        ggplot(aes(
            x = num_body,
            y = mean,
            color = factor(body_units),
            group = interaction(body_units, body_dropout)
        )) +
        geom_point() +
        geom_line(alpha = 0.35) +
        facet_wrap(~body_dropout) +
        labs(
            x = "num_body (profundidad)",
            y = "RMSE CV (media)",
            color = "body_units",
            title = "Tendencias en validación cruzada: profundidad / ancho / dropout"
        )
    print(p)
} else {
    cat(
        "Todavía no hay resultados en caché. Renderiza una vez para calcularlos.\n"
    )
}
```

::: {.notes}
Decir:
- Si las líneas se aplanan al aumentar num_body, más profundidad no aporta, elige el modelo menos profundo.
- Cada faceta es un valor de dropout: busca un panel donde la diferencia entre opciones de ancho sea pequeña. Esa es una zona robusta.

Transición: Confirmación final en test.
:::

---

## Resultados (Diamantes): Test — Tabla

```{r}
#| label: diamonds-results-test-table-es
#| echo: false
#| results: asis
#| message: false
#| warning: false
diamonds_final_rds <- file.path("results", "diamonds_final.rds")

if (file.exists(diamonds_final_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
    })
    final <- readRDS(diamonds_final_rds)
    metrics_tbl <- final$metrics |>
        dplyr::select(.metric, .estimate) |>
        dplyr::mutate(.estimate = round(.estimate, 4))
    cat(knitr::kable(metrics_tbl, format = "html"), "\n\n")
} else {
    cat("Todavía no hay resultados finales en caché.\n")
}
```

::: {.notes}
Decir:
- El test es solo para reportar. La arquitectura se eligió con CV, esto es confirmación.
- Si el test empeora mucho respecto a CV, sospecha data leakage o una arquitectura inestable.
:::

---

## Resultados (Diamantes): Test — Gráfico

```{r}
#| label: diamonds-results-test-plot-es
#| echo: false
#| message: false
#| warning: false
diamonds_final_rds <- file.path("results", "diamonds_final.rds")

if (file.exists(diamonds_final_rds)) {
    suppressPackageStartupMessages({
        library(ggplot2)
    })
    final <- readRDS(diamonds_final_rds)
    p_test <- final$preds |>
        ggplot(aes(x = price, y = .pred)) +
        geom_point(alpha = 0.25) +
        geom_abline(slope = 1, intercept = 0, linewidth = 0.8) +
        coord_fixed() +
        labs(
            x = "Valor real (price)",
            y = "Predicho",
            title = "Modelo final en test"
        )
    print(p_test)
} else {
    cat("Todavía no hay resultados finales en caché.\n")
}
```

::: {.notes}
Decir:
- Puntos cerca de la diagonal son buenas predicciones. Desviación sistemática hacia abajo en precios altos significa que el modelo subestima diamantes caros, esperable, ya que son raros en el entrenamiento.

Transición: Antes de pasar al API funcional, una herramienta que ahorra mucho tiempo de depuración.
:::

---

## Depuración: `compile_keras_grid()` {.compact .table-compact}

```r
prep_rec <- prep(diamonds_rec, training = diamonds_train)
juiced <- juice(prep_rec)
x <- juiced |> select(-price)
y <- juiced$price
modified_grid <- tibble(
    num_body = c(3, 6),
    body_units = c(-10, 128),
    body_dropout = c(0.2, 0.2)
)

dbg <- compile_keras_grid(spec, modified_grid, x, y)
dbg
```

```{r}
#| label: diamonds-debug-es
#| echo: false
#| message: false
#| warning: false
suppressPackageStartupMessages({
    library(dplyr)
    library(tibble)
    library(purrr)
    library(tidymodels)
    library(kerasnip)
    library(keras3)
})

if (
    !exists("diamond_mlp", mode = "function") ||
        !kerasnip::model_exists("diamond_mlp")
) {
    input_block <- function(model, input_shape) {
        keras_model_sequential(input_shape = input_shape)
    }
    dense_block <- function(model, units = 128, dropout = 0.0) {
        model |>
            layer_dense(units = units, activation = "relu") |>
            layer_dropout(rate = dropout)
    }
    output_block <- function(model) model |> layer_dense(units = 1)
    create_keras_sequential_spec(
        model_name = "diamond_mlp",
        layer_blocks = list(
            input = input_block,
            body = dense_block,
            output = output_block
        ),
        mode = "regression"
    )
}

prep_rec <- prep(diamonds_rec, training = diamonds_train)
juiced <- juice(prep_rec)
x <- juiced |> select(-price)
y <- juiced$price

test_grid <- tibble(
    num_body = c(3, 6),
    body_units = c(-10, 128),
    body_dropout = c(0.2, 0.2)
)

dbg <- compile_keras_grid(
    spec = diamond_mlp(
        num_body = tune(),
        body_units = tune(),
        body_dropout = tune(),
        fit_epochs = 5
    ) |>
        set_engine("keras"),
    grid = test_grid,
    x = x,
    y = y
)

dbg
```

::: {.notes}
Decir:
- Antes de lanzar tune_grid con 20 configuraciones, ejecuta esto con una para confirmar que el modelo construye correctamente.
- La columna error indica qué filas del grid fallaron y por qué, detecta esto antes de gastar cómputo.
- extract_valid_grid() filtra los builds correctos; inform_errors() resume los fallos en todo el grid.
- Regla: siempre compila antes de ajustar.

Transición: El mismo patrón de workflow, ahora con un grafo no lineal. API funcional.
:::

---

## API Funcional: Bloques + Grafo

::: {.columns}
::: {.column width="50%"}
**Los bloques reciben tensores, no modelos**

```r
input_block <- function(input_shape) {
  layer_input(shape = input_shape,
              name = "features")
}

dense_block <- function(tensor,
                        units = 64,
                        dropout = 0.0) {
  tensor |>
    layer_dense(units = units,
                activation = "relu") |>
    layer_dropout(rate = dropout)
}

concat_block <- function(input_a, input_b) {
  layer_concatenate(list(input_a, input_b))
}

output_block <- function(tensor, num_classes) {
  tensor |>
    layer_dense(units = num_classes,
                activation = "softmax")
}
```
:::

::: {.column width="50%"}
**`inp_spec()` conecta salidas con entradas por nombre**

```r
create_keras_functional_spec(
  model_name = "attrition_towers",
  layer_blocks = list(
    main_input = input_block,
    tower_a    = inp_spec(
                   dense_block,
                   "main_input"),
    tower_b    = inp_spec(
                   dense_block,
                   "main_input"),
    joined     = inp_spec(
                   concat_block,
                   c(input_a = "tower_a",
                     input_b = "tower_b")),
    output     = inp_spec(
                   output_block,
                   "joined")
  ),
  mode = "classification"
)
```
:::
:::

::: {.notes}
Decir:
- Secuencial: los bloques reciben un modelo y apilan capas. Funcional: los bloques reciben tensores y devuelven tensores. Esa es la única diferencia estructural.
- inp_spec(func, inputs) solo etiqueta qué salida upstream alimenta qué argumento, como cables de parcheo en un sintetizador modular.
- Este grafo: una entrada se divide en dos torres en paralelo, se concatenan, luego la salida. No lineal, pero el workflow es completamente idéntico.
- Usa secuencial para una cadena recta. Usa funcional cuando necesites ramas, uniones o skip connections.
- tower_a_units y tower_b_units se generan automáticamente con la misma convención de prefijos que antes.

Transición: El workflow, reconocerás cada línea.
:::

---

## Workflow de Rotación

```r
library(tidymodels);  library(modeldata)
data(attrition)

attr_split <- initial_split(attrition, strata = Attrition)
attr_train <- training(attr_split)

attr_rec <- recipe(Attrition ~ ., data = attr_train) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

attr_spec <- attrition_towers(
  tower_a_units = tune(),
  tower_b_units = tune(),
  fit_epochs    = 20
) |> set_engine("keras")

attr_wf <- workflow() |> add_recipe(attr_rec) |> add_model(attr_spec)

params <- extract_parameter_set_dials(attr_wf) |>
  update(tower_a_units = dials::hidden_units(c(16, 128)),
         tower_b_units = dials::hidden_units(c(16, 128)))

res <- tune_grid(attr_wf,
                 resamples = vfold_cv(attr_train, v = 3, strata = Attrition),
                 grid      = grid_latin_hypercube(params, size = 12))
```

::: {.notes}
Decir:
- Patrón idéntico al ejemplo secuencial. Ese es el punto de kerasnip, la complejidad del grafo queda completamente oculta.
- Asegúrate de que el outcome sea factor con niveles estables.

Transición: Resultados.
:::

---

## Resultados (Rotación): CV — Tabla

```{r}
#| label: attrition-results-cv-table-es
#| echo: false
#| results: asis
#| message: false
#| warning: false
attrition_rds <- file.path("results", "attrition_res.rds")

if (file.exists(attrition_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
        library(tune)
    })
    attr_res <- readRDS(attrition_rds)
    best_tbl <- show_best(attr_res, metric = "roc_auc", n = 5)
    best_tbl <- best_tbl |>
        dplyr::select(tower_a_units, tower_b_units, mean, std_err) |>
        dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 4)))
    cat(knitr::kable(best_tbl, format = "html"), "\n\n")
} else {
    cat(
        "Todavía no hay resultados en caché. Renderiza una vez para calcularlos.\n"
    )
}
```

::: {.notes}
Mantener en 30 segundos y pasar al mapa de calor.

Decir:
- Misma lectura que antes: ROC AUC medio más std_err. Seleccionamos con CV.
- Elige la arquitectura más simple dentro de 1 std_err del mejor.
:::

---

## Resultados (Rotación): CV — Gráfico

```{r}
#| label: attrition-results-cv-plot-es
#| echo: false
#| message: false
#| warning: false
attrition_rds <- file.path("results", "attrition_res.rds")

if (file.exists(attrition_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
        library(ggplot2)
        library(tune)
    })
    attr_res <- readRDS(attrition_rds)
    p_heat <- collect_metrics(attr_res) |>
        filter(.metric == "roc_auc") |>
        ggplot(aes(
            x = factor(tower_a_units),
            y = factor(tower_b_units),
            fill = mean
        )) +
        geom_tile() +
        scale_fill_gradient(low = "#f0f1f2", high = "#BF281B") +
        labs(
            x = "tower_a_units",
            y = "tower_b_units",
            fill = "media roc_auc",
            title = "Media roc_auc (CV)"
        )
    print(p_heat)
} else {
    cat(
        "Todavía no hay resultados en caché. Renderiza una vez para calcularlos.\n"
    )
}
```

::: {.notes}
Decir:
- Un mapa plano significa que la arquitectura es robusta, el tamaño de las torres no importa mucho, elige el más pequeño.
- Un pico claro significa que el modelo es sensible, vale la pena acotar la búsqueda alrededor de esa zona en una segunda pasada.

Transición: Confirmación final en test.
:::

---

## Resultados (Rotación): Test — Tabla

```{r}
#| label: attrition-results-test-table-es
#| echo: false
#| results: asis
#| message: false
#| warning: false
attrition_final_rds <- file.path("results", "attrition_final.rds")

if (file.exists(attrition_final_rds)) {
    suppressPackageStartupMessages({
        library(dplyr)
    })
    final <- readRDS(attrition_final_rds)
    metrics_tbl <- final$metrics |>
        dplyr::select(.metric, .estimate) |>
        dplyr::mutate(.estimate = round(.estimate, 4))
    cat(knitr::kable(metrics_tbl, format = "html"), "\n\n")
} else {
    cat("Todavía no hay resultados finales en caché.\n")
}
```

::: {.notes}
Decir:
- En conjuntos desbalanceados como attrition, el ROC AUC es más informativo que el accuracy, el accuracy puede verse bien simplemente prediciendo la clase mayoritaria.
- De nuevo: el test es para reportar, no para decidir.
:::

---

## Resultados (Rotación): Test — Gráfico

```{r}
#| label: attrition-results-test-plot-es
#| echo: false
#| message: false
#| warning: false
attrition_final_rds <- file.path("results", "attrition_final.rds")

if (file.exists(attrition_final_rds)) {
    suppressPackageStartupMessages({
        library(ggplot2)
    })
    final <- readRDS(attrition_final_rds)
    if (!is.null(final$roc)) {
        p_roc <- ggplot(final$roc, aes(x = 1 - specificity, y = sensitivity)) +
            geom_line(linewidth = 1) +
            geom_abline(linetype = "dashed", alpha = 0.5) +
            coord_equal() +
            labs(
                x = "Tasa de falsos positivos",
                y = "Tasa de verdaderos positivos",
                title = "Curva ROC (test)"
            )
        print(p_roc)
    } else {
        cat("No hay curva ROC disponible.\n")
    }
} else {
    cat("Todavía no hay resultados finales en caché.\n")
}
```

::: {.notes}
Decir:
- Cuanto más lejos de la diagonal, mejor la discriminación.
- Cuando dos arquitecturas dan curvas similares, elige siempre la más simple.

Transición: Cerramos con el resumen práctico.
:::

---

## Consejos Prácticos y Valoración

::: {.columns}
::: {.column width="50%"}
**Para evitar problemas**

- Ejecuta siempre `compile_keras_grid()` antes de ajustar
- Bloques pequeños y testeables, aísla errores con 2–3 bloques
- Ajusta estructura (profundidad/ancho) primero; hiperparámetros de entrenamiento después
- Mantén el espacio de búsqueda razonable

:::

::: {.column width="50%"}
**Por qué vale la pena**

1. **Paridad de workflow** — `recipes`, `workflows`, `tune` funcionan igual que con `glmnet`
2. **Reproducibilidad** — la topología es código; controla tu arquitectura con git
3. **Extensibilidad** — envuelve cualquier capa Keras (Transformers, RNNs) en un bloque

<br/>

:::
:::

::: {.notes}
Objetivo: que la gente salga con una respuesta clara a "¿debería probar esto?"

Decir:
- La curva de aprendizaje está en escribir buenos bloques. Después de eso, todo es tidymodels estándar.
- Si ya usas tidymodels, lo único nuevo es definir bloques y llamar a create_keras_*_spec().
- El paquete está en GitHub, contribuciones y feedback bienvenidos.

Dejar de compartir pantalla y abrir turno de preguntas.
:::
---

## Recursos

- 📦 [github.com/davidrsch/kerasnip](https://github.com/davidrsch/kerasnip)
- 📑 [davidrsch.github.io/kerasnip](https://davidrsch.github.io/kerasnip/)
- 📑 [https://davidrsch.github.io/kerasnip_presentations/presentations/madrid-2-2026/es/)

:::
